---
title: "Module 6 - Exercises"
subtitle: "STT-1100 Introduction to Data Science"
format:
  html:
    toc: true
    toc-title: Plan
    embed-resources: true
    link-external-newwindow: true
    css: [../css/base_css.css]
editor: visual
---

::: {.callout-important}
## Module workspace

<a class="btn btn-primary" href="../../downloads/donnees/stt1100-module-06-en.zip" download>Download the ready-to-open workspace (.zip)</a>

Unpack the file, then open the `.Rproj` project. The paths used in the exercises will work without moving the data.
:::


## Consolidation Exercises

These exercises are independent from the adventure and challenge. They consolidate the technical moves from Module 6: joining tables, checking keys, finding unmatched rows, understanding cardinalities and writing constructive reviews.

::: {.callout-note}
## A real Québec relational catalogue

The four tables come from the [official Données Québec API](https://www.donneesquebec.ca/recherche/api/3/action/package_search). The extract contains 142 organizations, 312 datasets, 3,143 resources and 2,101 dataset-tag associations. To prevent one organization from dominating the exercises, the script retains at most eight datasets among the 1,000 most recently modified for each organization. The result includes ministries and cities from several regions, including Laval, Lévis, Longueuil, Montréal, Québec City, Rimouski, Rouyn-Noranda, Saguenay, Sherbrooke and Trois-Rivières. The script `scripts/build_module06_real_data.R` rebuilds the extract.
:::

```{r setup, warning=FALSE, message=FALSE}
library(tidyverse)
```

## Block A - Understand Join Keys

### Exercise 1 - Import Organizations and Datasets

Import `organisations_donnees_quebec.csv` and `jeux_donnees_quebec.csv`, then inspect their dimensions and keys. The official French field names are retained.

::: {.callout-note collapse="true"}
### Solution

```{r}
organizations <- read_csv(
  "data/organisations_donnees_quebec.csv",
  show_col_types = FALSE
)

datasets <- read_csv(
  "data/jeux_donnees_quebec.csv",
  show_col_types = FALSE
)

glimpse(organizations)
glimpse(datasets)

tibble(
  table = c("organizations", "datasets"),
  rows = c(nrow(organizations), nrow(datasets)),
  distinct_keys = c(
    n_distinct(organizations$organisation_id),
    n_distinct(datasets$jeu_id)
  )
)
```
:::

### Exercise 2 - Add the Organization Name

Use `left_join()` to add each organization's name and total catalogue dataset count to each dataset in the extract.

::: {.callout-note collapse="true"}
### Solution

```{r}
datasets_detail <- datasets |>
  left_join(organizations, by = "organisation_id")

datasets_detail |>
  select(
    jeu_id,
    titre,
    organisation_nom,
    licence,
    nb_ressources,
    nb_jeux_catalogue
  ) |>
  slice_head(n = 10)
```
:::

### Exercise 3 - Find Datasets Without a Known Organization

Use `anti_join()` to check whether any dataset has an organization key absent from the organization table.

::: {.callout-note collapse="true"}
### Solution

```{r}
datasets |>
  anti_join(organizations, by = "organisation_id")
```

Returning zero rows is useful evidence: every foreign key in this extract has a match.
:::

### Exercise 4 - Identify Represented Organizations

Use `semi_join()` to retain only organizations with at least one dataset in the balanced extract.

::: {.callout-note collapse="true"}
### Solution

```{r}
represented_organizations <- organizations |>
  semi_join(datasets, by = "organisation_id") |>
  arrange(organisation_nom)

represented_organizations
```
:::

## Block B - One-to-Many Relationships

### Exercise 5 - Summarize Datasets by Organization

Calculate the number of datasets retained in the extract and the total catalogue count for each represented organization.

::: {.callout-note collapse="true"}
### Solution

```{r}
organization_summary <- datasets_detail |>
  group_by(organisation_id, organisation_nom, nb_jeux_catalogue) |>
  summarise(
    extract_datasets = n(),
    extract_resources = sum(nb_ressources),
    .groups = "drop"
  ) |>
  arrange(desc(nb_jeux_catalogue))

organization_summary
```
:::

### Exercise 6 - Visualize the Largest Organizations

Produce a graph of the ten organizations with the most datasets in the complete catalogue.

::: {.callout-note collapse="true"}
### Solution

```{r}
organizations |>
  slice_max(nb_jeux_catalogue, n = 10, with_ties = FALSE) |>
  ggplot(
    aes(
      x = fct_reorder(organisation_nom, nb_jeux_catalogue),
      y = nb_jeux_catalogue
    )
  ) +
  geom_col(fill = "#2c7fb8") +
  coord_flip() +
  labs(
    title = "Organizations with the most Données Québec datasets",
    x = NULL,
    y = "Number of catalogue datasets"
  )
```
:::

### Exercise 7 - Join Resources to Datasets

Import `ressources_donnees_quebec.csv`, add the dataset title to each resource and count formats.

::: {.callout-note collapse="true"}
### Solution

```{r}
resources <- read_csv(
  "data/ressources_donnees_quebec.csv",
  show_col_types = FALSE
)

resources_detail <- resources |>
  left_join(
    datasets |> select(jeu_id, titre, organisation_id),
    by = "jeu_id",
    relationship = "many-to-one"
  )

resources_detail |>
  count(format, sort = TRUE) |>
  slice_head(n = 15)
```
:::

### Exercise 8 - Check Dataset-Resource Matches

Find datasets without resources and resources without a known dataset.

::: {.callout-note collapse="true"}
### Solution

```{r}
datasets_without_resource <- datasets |>
  anti_join(resources, by = "jeu_id")

resources_without_dataset <- resources |>
  anti_join(datasets, by = "jeu_id")

datasets_without_resource
resources_without_dataset
```
:::

## Block C - Many-to-Many Relationships and Review

### Exercise 9 - Add Tags

Import `etiquettes_donnees_quebec.csv`, then add the dataset title and organization to each dataset-tag association.

::: {.callout-note collapse="true"}
### Solution

```{r}
tags <- read_csv(
  "data/etiquettes_donnees_quebec.csv",
  show_col_types = FALSE
)

tags_detail <- tags |>
  left_join(
    datasets_detail |> select(jeu_id, titre, organisation_nom),
    by = "jeu_id",
    relationship = "many-to-one"
  )

tags_detail |>
  slice_head(n = 10)
```
:::

### Exercise 10 - Summarize Tags

Find the fifteen tags associated with the largest number of distinct datasets.

::: {.callout-note collapse="true"}
### Solution

```{r}
tag_summary <- tags_detail |>
  summarise(
    dataset_count = n_distinct(jeu_id),
    .by = etiquette
  ) |>
  arrange(desc(dataset_count)) |>
  slice_head(n = 15)

tag_summary
```
:::

### Exercise 11 - Observe Row Multiplication

Join resources and then tags to datasets. Compare row counts before and after and explain why the second join is many-to-many.

::: {.callout-note collapse="true"}
### Solution

```{r}
datasets_resources <- datasets |>
  left_join(
    resources,
    by = "jeu_id",
    relationship = "one-to-many"
  )

flattened_catalogue <- datasets_resources |>
  left_join(
    tags,
    by = "jeu_id",
    relationship = "many-to-many"
  )

tibble(
  table = c("datasets", "datasets_resources", "flattened_catalogue"),
  rows = c(
    nrow(datasets),
    nrow(datasets_resources),
    nrow(flattened_catalogue)
  )
)
```

A dataset may have several resources and several tags. Resource-tag combinations therefore multiply rows without creating new datasets.
:::

### Exercise 12 - Write a Useful Issue

Find a resource with a missing format, then draft an issue with three sections: finding, suggestion and justification.

::: {.callout-note collapse="true"}
### Example Answer

```markdown
## Finding

The resource record does not specify a format in the catalogue metadata.

## Suggestion

Populate the format field with a normalized value matching the published file or service.

## Why this is useful

The format helps users select tools and filter resources before downloading them.
```
:::

## Case Studies

### Case Study 1 - Relational Audit of Données Québec

Use the organization, dataset, resource and tag tables.

Complete the following tasks:

1. check the uniqueness of the three primary keys;
2. join datasets to organizations;
3. identify unmatched foreign keys;
4. find datasets without resources and without tags;
5. produce a summary by organization;
6. write a cautious conclusion about relational integrity.

::: {.callout-note collapse="true"}
### Solution

```{r}
key_check <- tibble(
  table = c("organizations", "datasets", "resources"),
  rows = c(nrow(organizations), nrow(datasets), nrow(resources)),
  distinct_keys = c(
    n_distinct(organizations$organisation_id),
    n_distinct(datasets$jeu_id),
    n_distinct(resources$ressource_id)
  )
)

datasets_organizations <- datasets |>
  left_join(organizations, by = "organisation_id")

datasets_without_organization <- datasets |>
  anti_join(organizations, by = "organisation_id")

resources_without_dataset <- resources |>
  anti_join(datasets, by = "jeu_id")

datasets_without_resource <- datasets |>
  anti_join(resources, by = "jeu_id")

datasets_without_tag <- datasets |>
  anti_join(tags, by = "jeu_id")

audit_summary <- datasets_organizations |>
  group_by(organisation_nom) |>
  summarise(
    dataset_count = n(),
    resource_count = sum(nb_ressources),
    datasets_without_tag = sum(nb_etiquettes == 0),
    .groups = "drop"
  ) |>
  arrange(desc(resource_count))

key_check
datasets_without_organization
resources_without_dataset
datasets_without_resource
datasets_without_tag
audit_summary
```

A cautious conclusion may note the absence of orphan keys in the supplied extract. It must also state that the extract is balanced by organization and represents neither the full catalogue nor its real distribution.
:::

### Case Study 2 - Review Resource Metadata

Use the same tables to conduct a constructive review of resource documentation.

Complete the following tasks:

1. join resources to datasets and organizations;
2. identify missing formats;
3. calculate the share of resources without a declared size;
4. summarize formats by organization;
5. identify datasets without tags;
6. draft a precise issue for an observed problem.

::: {.callout-note collapse="true"}
### Solution

```{r}
catalogue_resources <- resources |>
  left_join(
    datasets |> select(jeu_id, titre, organisation_id),
    by = "jeu_id",
    relationship = "many-to-one"
  ) |>
  left_join(
    organizations |> select(organisation_id, organisation_nom),
    by = "organisation_id",
    relationship = "many-to-one"
  )

missing_formats <- catalogue_resources |>
  filter(is.na(format) | str_squish(format) == "")

missing_size_share <- catalogue_resources |>
  summarise(
    resource_count = n(),
    resources_without_size = sum(is.na(taille_octets)),
    share_without_size = mean(is.na(taille_octets))
  )

formats_by_organization <- catalogue_resources |>
  count(organisation_nom, format, sort = TRUE)

datasets_without_tag <- datasets |>
  anti_join(tags, by = "jeu_id")

missing_formats
missing_size_share
formats_by_organization
datasets_without_tag
```

A useful issue identifies the resource, describes the missing field and proposes a verifiable correction. A missing size is not automatically an error, especially for a service or interactive map resource.
:::
