---
title: "Module 6 - Exercises"
subtitle: "STT-1100 Introduction to Data Science"
format:
  html:
    toc: true
    toc-title: Plan
    embed-resources: true
    link-external-newwindow: true
    css: [../css/base_css.css]
editor: visual
---

# Consolidation Exercises

These exercises are independent from the adventure and challenge. They consolidate the technical moves from module 6: join several tables, check keys, find unmatched rows, summarize collaborative activity and write constructive review comments.

The data used here are fictitious and do not represent any real GitHub repository.

```{r setup, warning=FALSE, message=FALSE}
library(tidyverse)
```

# Readings to Review

- [GitHub Docs - Hello World](https://docs.github.com/en/get-started/quickstart/hello-world){target="_blank"}
- [GitHub Docs - About issues](https://docs.github.com/en/issues/tracking-your-work-with-issues/about-issues){target="_blank"}
- [GitHub Docs - Creating an issue](https://docs.github.com/en/issues/tracking-your-work-with-issues/using-issues/creating-an-issue){target="_blank"}
- [GitHub Docs - About pull requests](https://docs.github.com/en/pull-requests){target="_blank"}
- [GitHub Docs - About merge conflicts](https://docs.github.com/en/pull-requests/collaborating-with-pull-requests/addressing-merge-conflicts/about-merge-conflicts){target="_blank"}
- [Quarto Docs - Inline code](https://quarto.org/docs/computations/inline-code.html){target="_blank"}
- [Quarto Docs - Execution options](https://quarto.org/docs/computations/execution-options.html){target="_blank"}
- [R for Data Science - Joins](https://r4ds.hadley.nz/joins.html){target="_blank"}

After the readings, also complete the [formative mini-test](mini_test.qmd). It is not graded.

# Block A - Understand Join Keys

## Exercise 1 - Import Collaboration Tables

Import `fictitious_team_members.csv` and `fictitious_git_contributions.csv`, then inspect their dimensions.

::: {.callout-note collapse="true"}
## Solution

```{r}
members <- read_csv(
  "data/fictitious_team_members.csv",
  show_col_types = FALSE
)

contributions <- read_csv(
  "data/fictitious_git_contributions.csv",
  show_col_types = FALSE
)

glimpse(members)
glimpse(contributions)

tibble(
  table = c("members", "contributions"),
  rows = c(nrow(members), nrow(contributions)),
  columns = c(ncol(members), ncol(contributions))
)
```
:::

## Exercise 2 - Add Member Information

Use `left_join()` to add the name, role and team to each contribution.

::: {.callout-note collapse="true"}
## Solution

```{r}
contribution_details <- contributions |>
  left_join(members, by = "member_id")

contribution_details |>
  select(commit_id, member_id, name, role, team, contribution_type, file) |>
  arrange(commit_id)
```
:::

## Exercise 3 - Find Contributions Without a Known Member

Use `anti_join()` to find contributions whose `member_id` does not appear in the member table.

::: {.callout-note collapse="true"}
## Solution

```{r}
contributions |>
  anti_join(members, by = "member_id")
```

This check is useful before summarizing work by person. A missing key may indicate a data-entry error or an undocumented member.
:::

## Exercise 4 - Identify Members Who Contributed

Use `semi_join()` to get the list of members who have at least one contribution.

::: {.callout-note collapse="true"}
## Solution

```{r}
members |>
  semi_join(contributions, by = "member_id") |>
  arrange(team, role)
```
:::

# Block B - Summarize Collaborative Work

## Exercise 5 - Summarize Contributions by Member

Create a table giving, for each known member, the number of commits and the total number of changed lines.

::: {.callout-note collapse="true"}
## Solution

```{r}
member_summary <- contribution_details |>
  filter(!is.na(name)) |>
  group_by(team, name, role) |>
  summarise(
    n_commits = n(),
    changed_lines = sum(changed_lines),
    .groups = "drop"
  ) |>
  arrange(team, desc(n_commits))

member_summary
```
:::

## Exercise 6 - Visualize Team Activity

Produce a chart of the number of commits by role.

::: {.callout-note collapse="true"}
## Solution

```{r}
member_summary |>
  ggplot(aes(x = role, y = n_commits, fill = team)) +
  geom_col(position = "dodge") +
  coord_flip() +
  labs(
    title = "Number of commits by role",
    x = "Role",
    y = "Number of commits",
    fill = "Team"
  )
```
:::

## Exercise 7 - Check Expected Files

Import `fictitious_report_files.csv`. For each team, calculate the number of expected files present and missing.

::: {.callout-note collapse="true"}
## Solution

```{r}
files <- read_csv(
  "data/fictitious_report_files.csv",
  show_col_types = FALSE
)

file_summary <- files |>
  mutate(is_present = present == "yes") |>
  group_by(team) |>
  summarise(
    files_present = sum(is_present),
    files_missing = sum(!is_present),
    total_size_kb = sum(size_kb),
    .groups = "drop"
  )

file_summary
```
:::

## Exercise 8 - Find Teams Missing an HTML Report

Filter rows where `expected_file == "rapport.html"` and the file is absent.

::: {.callout-note collapse="true"}
## Solution

```{r}
files |>
  filter(expected_file == "rapport.html", present == "no")
```
:::

# Block C - Issues and Constructive Review

## Exercise 9 - Join Issues to Members

Import `fictitious_review_issues.csv`, then add the name and team of the author of each issue.

::: {.callout-note collapse="true"}
## Solution

```{r}
issues <- read_csv(
  "data/fictitious_review_issues.csv",
  show_col_types = FALSE
)

issue_details <- issues |>
  left_join(members, by = c("author_id" = "member_id"))

issue_details |>
  select(issue_id, reviewed_team, author_id, name, team, dimension, status, priority)
```
:::

## Exercise 10 - Summarize Issues by Reviewed Team

Calculate the number of open, closed and high-priority issues for each reviewed team.

::: {.callout-note collapse="true"}
## Solution

```{r}
issue_details |>
  group_by(reviewed_team) |>
  summarise(
    n_issues = n(),
    n_open = sum(status == "open"),
    n_closed = sum(status == "closed"),
    n_high_priority = sum(priority == "high"),
    .groups = "drop"
  ) |>
  arrange(desc(n_open))
```
:::

## Exercise 11 - Find Issues Without a Known Team

Teams A and B are in the member table. Use a join or equivalent logic to identify issues targeting a team absent from the known team list.

::: {.callout-note collapse="true"}
## Solution

```{r}
known_teams <- members |>
  distinct(team)

issues |>
  anti_join(known_teams, by = c("reviewed_team" = "team"))
```
:::

## Exercise 12 - Write a Useful Issue

Choose one open issue and rewrite it with three sections: successful aspect, suggestion, justification.

::: {.callout-note collapse="true"}
## Example Answer

```markdown
## Successful aspect

The report clearly lists the packages used and the main preparation steps.

## Suggestion

Add one sentence explaining the join key used between contributions and members.

## Why this suggestion is useful

The review becomes easier to reproduce because another team can understand why `member_id` is the correct key.
```
:::

# Case Studies

## Case Study 1 - Fictitious Project Repository Audit

Use `fictitious_team_members.csv`, `fictitious_git_contributions.csv`, `fictitious_review_issues.csv` and `fictitious_report_files.csv`.

Complete the following tasks:

1. join contributions to members;
2. find contributions without a known member;
3. produce a summary by team;
4. join issues to authors;
5. identify teams missing `rapport.html`;
6. write a cautious conclusion about repository reproducibility.

::: {.callout-note collapse="true"}
## Solution

```{r}
audit_contributions <- contributions |>
  left_join(members, by = "member_id")

contributions_without_member <- contributions |>
  anti_join(members, by = "member_id")

team_summary <- audit_contributions |>
  filter(!is.na(team)) |>
  group_by(team) |>
  summarise(
    n_commits = n(),
    n_active_members = n_distinct(member_id),
    changed_lines = sum(changed_lines),
    .groups = "drop"
  )

audit_issues <- issues |>
  left_join(members, by = c("author_id" = "member_id"))

missing_reports <- files |>
  filter(expected_file == "rapport.html", present == "no")

contributions_without_member
team_summary
audit_issues
missing_reports
```

A cautious conclusion could state that Team B appears to have its data trace present, but an HTML report is missing. This observation is not enough to judge the whole project; the next step would be to try rendering `rapport.qmd`.
:::

## Case Study 2 - Fictitious Data Portal Review

Use `fictitious_data_portal.csv`, `fictitious_producers.csv` and `fictitious_review_comments.csv`.

Complete the following tasks:

1. join datasets to producers;
2. find datasets without a known producer;
3. join comments to datasets;
4. count comments by dimension and severity;
5. identify datasets with an unknown license;
6. write a constructive review issue for a priority problem.

::: {.callout-note collapse="true"}
## Solution

```{r}
portal <- read_csv(
  "data/fictitious_data_portal.csv",
  show_col_types = FALSE
)

producers <- read_csv(
  "data/fictitious_producers.csv",
  show_col_types = FALSE
)

comments <- read_csv(
  "data/fictitious_review_comments.csv",
  show_col_types = FALSE
)

portal_producers <- portal |>
  left_join(producers, by = "producer_id")

datasets_without_producer <- portal |>
  anti_join(producers, by = "producer_id")

comment_details <- comments |>
  left_join(portal, by = "dataset_id")

comment_summary <- comment_details |>
  count(dimension, severity, sort = TRUE)

unknown_licenses <- portal_producers |>
  filter(license == "Unknown" | is.na(organization))

datasets_without_producer
comment_summary
unknown_licenses
```

A priority issue could ask the team to clarify the license for the `Road work` dataset before publishing. The justification is that data reuse directly depends on the stated license.
:::
