---
title: "Module 5 - Exercises"
subtitle: "STT-1100 Introduction to Data Science"
format:
  html:
    toc: true
    toc-title: Plan
    embed-resources: true
    link-external-newwindow: true
    css: [../css/base_css.css]
editor: visual
---

## Consolidation Exercises

These exercises are independent from the adventure and challenge. They consolidate the technical moves from Module 5: manipulating dates, summarizing groups, visualizing associations, calculating simple correlations and writing cautious conclusions.

::: {.callout-note}
## Three real Québec sources

The exercises use [Laval bicycle counts](https://www.donneesquebec.ca/recherche/dataset/comptages-velos), [hourly Québec air-quality measurements](https://www.donneesquebec.ca/recherche/dataset/rsqaq-donnees-horaires-continues) from the Québec - Vieux-Limoilou station and [Gatineau traffic volumes](https://www.donneesquebec.ca/recherche/dataset/debits-de-circulation). These official sources are released under the CC BY 4.0 licence. The local files are reproducible extracts prepared by `scripts/build_module05_real_data.R`.
:::

```{r setup, warning=FALSE, message=FALSE}
library(tidyverse)
library(lubridate)
```

## Readings to Review

- [R for Data Science - Exploratory data analysis](https://r4ds.hadley.nz/EDA.html){target="_blank"}
- [R for Data Science - Dates and times](https://r4ds.hadley.nz/datetimes.html){target="_blank"}
- [R for Data Science - Data visualization](https://r4ds.hadley.nz/data-visualize.html){target="_blank"}
- [R for Data Science - Missing values](https://r4ds.hadley.nz/missing-values.html){target="_blank"}
- [Introduction to Modern Statistics - Exploring numerical data](https://openintrostat.github.io/ims/explore-numerical){target="_blank"}
- [Introduction to Modern Statistics - Applications: Explore](https://openintrostat.github.io/ims/explore-applications){target="_blank"}

After the readings, also complete the [formative mini-test](mini_test.qmd). It is not graded.

## Block A - Dates, Structure and Missing Values

The file `data/comptages_velos_laval_2016_06.csv` contains counts recorded every 15 minutes at Laval's Chevillon loop in June 2016. The source's French field names are retained.

### Exercise 1 - Import and Inspect

Import the file, display its structure and check its dimensions.

::: {.callout-note collapse="true"}
### Solution

```{r}
bikes <- read_csv(
  "data/comptages_velos_laval_2016_06.csv",
  show_col_types = FALSE
)

glimpse(bikes)

tibble(
  rows = nrow(bikes),
  columns = ncol(bikes)
)
```
:::

### Exercise 2 - Prepare Temporal Variables

Combine `date` and `heure`, then create `weekday`, `month`, `hour_of_day` and `time_of_day`.

::: {.callout-note collapse="true"}
### Solution

```{r}
bikes_dates <- bikes |>
  mutate(
    date_time = ymd_hms(paste(date, heure)),
    weekday = wday(date_time, label = TRUE, abbr = FALSE),
    month = month(date_time, label = TRUE, abbr = FALSE),
    hour_of_day = hour(date_time),
    time_of_day = case_when(
      hour_of_day < 12 ~ "morning",
      hour_of_day < 17 ~ "afternoon",
      TRUE ~ "evening"
    )
  )

bikes_dates |>
  select(date_time, weekday, month, hour_of_day, time_of_day) |>
  slice_head(n = 8)
```
:::

### Exercise 3 - Check Missing Values and an Accounting Relationship

Count missing values in the three count columns. Also check whether `velos_total` equals the sum of inbound and outbound bicycles.

::: {.callout-note collapse="true"}
### Solution

```{r}
bikes_dates |>
  summarise(
    total_missing = sum(is.na(velos_total)),
    inbound_missing = sum(is.na(velos_entrants)),
    outbound_missing = sum(is.na(velos_sortants)),
    inconsistent_totals = sum(
      velos_total != velos_entrants + velos_sortants,
      na.rm = TRUE
    )
  )
```
:::

### Exercise 4 - Summarize by Time of Day

Calculate the number of observations and mean bicycle count by `time_of_day`.

::: {.callout-note collapse="true"}
### Solution

```{r}
time_summary <- bikes_dates |>
  group_by(time_of_day) |>
  summarise(
    n_observations = n(),
    mean_bikes = mean(velos_total, na.rm = TRUE),
    mean_inbound = mean(velos_entrants, na.rm = TRUE),
    mean_outbound = mean(velos_sortants, na.rm = TRUE),
    .groups = "drop"
  )

time_summary
```
:::

## Block B - Associations and Visualizations

### Exercise 5 - Visualize Mean Counts

Produce a column chart of the mean bicycle count by time of day.

::: {.callout-note collapse="true"}
### Solution

```{r}
time_summary |>
  ggplot(aes(x = time_of_day, y = mean_bikes)) +
  geom_col(fill = "#2c7fb8") +
  labs(
    title = "Mean count by time of day",
    x = "Time of day",
    y = "Mean number of bicycles per 15-minute interval"
  )
```
:::

### Exercise 6 - Compare Weekdays

For each weekday, calculate the number of observations and mean count.

::: {.callout-note collapse="true"}
### Solution

```{r}
summary_weekday <- bikes_dates |>
  group_by(weekday) |>
  summarise(
    n_observations = n(),
    mean_bicycles = mean(velos_total, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(desc(mean_bicycles))

summary_weekday
```
:::

### Exercise 7 - Calculate a Correlation Matrix

Calculate correlations among `velos_total`, `velos_entrants` and `velos_sortants`.

::: {.callout-note collapse="true"}
### Solution

```{r}
bikes_dates |>
  select(velos_total, velos_entrants, velos_sortants) |>
  cor(use = "complete.obs") |>
  round(2)
```

Here the correlation is affected by the accounting relationship between the total and the two directions. It is not evidence of causation.
:::

### Exercise 8 - Make a Scatterplot

Graph incoming and outgoing bicycles. Colour the points by time of day and add a trend line.

::: {.callout-note collapse="true"}
### Solution

```{r}
bikes_dates |>
  ggplot(aes(x = velos_entrants, y = velos_sortants)) +
  geom_point(aes(color = time_of_day), alpha = 0.45) +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  labs(
    title = "Incoming and outgoing bicycle counts in Laval",
    x = "Incoming bicycles",
    y = "Outgoing bicycles",
    color = "Time"
  )
```
:::

### Exercise 9 - Compare Weekdays and Weekends

Create a `day_type` variable, then compare mean counts between weekdays and weekends.

::: {.callout-note collapse="true"}
### Solution

```{r}
bikes_day_type <- bikes_dates |>
  mutate(
    day_type = if_else(
      wday(date_time) %in% c(1, 7),
      "weekend",
      "weekday"
    )
  )

bikes_day_type |>
  group_by(day_type) |>
  summarise(
    n_observations = n(),
    mean_bicycles = mean(velos_total, na.rm = TRUE),
    .groups = "drop"
  )
```
:::

## Block C - Cautious Interpretation

### Exercise 10 - Write a Descriptive Conclusion

Using one of the previous tables, write one descriptive sentence and one sentence stating a limitation.

::: {.callout-note collapse="true"}
### Example Answer

In this extract, the mean count is higher in the afternoon than in the morning or evening. The result describes only the Chevillon counter during June 2016 and cannot explain the cause of this difference.
:::

### Exercise 11 - Choose an Appropriate Visualization

Produce a graph of mean counts by hour of day.

::: {.callout-note collapse="true"}
### Possible Solution

```{r}
bikes_dates |>
  group_by(hour_of_day) |>
  summarise(
    mean_bicycles = mean(velos_total, na.rm = TRUE),
    .groups = "drop"
  ) |>
  ggplot(aes(x = hour_of_day, y = mean_bicycles)) +
  geom_line(linewidth = 1, color = "#2c7fb8") +
  geom_point(color = "#2c7fb8") +
  labs(
    title = "Hourly bicycle-count profile",
    x = "Hour",
    y = "Mean number of bicycles"
  )
```
:::

### Exercise 12 - Build a Daily Summary Table

Create a table by date containing the daily total, interval mean and largest 15-minute count.

::: {.callout-note collapse="true"}
### Solution

```{r}
bikes_dates |>
  group_by(date) |>
  summarise(
    daily_total = sum(velos_total, na.rm = TRUE),
    interval_mean = mean(velos_total, na.rm = TRUE),
    interval_maximum = max(velos_total, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(desc(daily_total))
```
:::

## Case Studies

### Case Study 1 - Air Quality at Québec - Vieux-Limoilou

The file `data/qualite_air_quebec_vieux_limoilou_2025_07.csv` contains 744 hourly measurements from July 2025 at the Québec - Vieux-Limoilou station. The variables are pollutant concentrations and may contain missing values.

Complete the following tasks:

1. import the file and transform `date_heure`;
2. create `date`, `hour_of_day` and `time_of_day`;
3. count missing values for each pollutant;
4. summarize mean concentrations by time of day;
5. calculate a correlation matrix among pollutants;
6. produce a time-series visualization of fine particles;
7. write a cautious conclusion.

::: {.callout-note collapse="true"}
### Solution

```{r}
quebec_air <- read_csv(
  "data/qualite_air_quebec_vieux_limoilou_2025_07.csv",
  show_col_types = FALSE
) |>
  mutate(
    date_heure = ymd_hms(date_heure),
    date = as_date(date_heure),
    hour_of_day = hour(date_heure),
    time_of_day = case_when(
      hour_of_day < 12 ~ "morning",
      hour_of_day < 17 ~ "afternoon",
      TRUE ~ "evening"
    )
  )

quebec_air |>
  summarise(
    across(
      carbone_noir:dioxyde_soufre,
      ~ sum(is.na(.x)),
      .names = "missing_{.col}"
    )
  )

quebec_air |>
  group_by(time_of_day) |>
  summarise(
    across(carbone_noir:dioxyde_soufre, ~ mean(.x, na.rm = TRUE)),
    .groups = "drop"
  )

quebec_air |>
  select(carbone_noir:dioxyde_soufre) |>
  cor(use = "complete.obs") |>
  round(2)

quebec_air |>
  ggplot(aes(x = date_heure, y = particules_fines)) +
  geom_line(color = "#7b3294", linewidth = 0.5, na.rm = TRUE) +
  labs(
    title = "Fine particles at Québec - Vieux-Limoilou",
    subtitle = "Hourly measurements, July 2025",
    x = NULL,
    y = "PM2.5"
  )
```

A cautious conclusion may describe hourly variation and observed associations. It must remain limited to one station and one month, without assigning the concentrations to a specific cause.
:::

### Case Study 2 - Traffic Volumes in Gatineau

The file `data/debits_circulation_gatineau_2016_2023.csv` is a teaching extract of traffic volumes published by the City of Gatineau for different intersections, approaches, directions and years. It retains 150 rows per year to keep calculations light, but is not a representative sample of the city.

Complete the following tasks:

1. import the file and inspect missing values;
2. create an estimate of the number of heavy vehicles;
3. summarize traffic volumes by sector and year;
4. calculate correlations among year, traffic volume, heavy-vehicle share and estimated heavy-vehicle count;
5. produce a useful visualization;
6. write a cautious conclusion.

::: {.callout-note collapse="true"}
### Solution

```{r}
traffic <- read_csv(
  "data/debits_circulation_gatineau_2016_2023.csv",
  show_col_types = FALSE
) |>
  mutate(
    estimated_heavy_vehicles =
      debit_total_24h * pourcentage_vehicules_lourds / 100
  )

traffic |>
  summarise(across(everything(), ~ sum(is.na(.x))))

traffic |>
  group_by(secteur, annee) |>
  summarise(
    n_observations = n(),
    median_volume = median(debit_total_24h, na.rm = TRUE),
    mean_heavy_percentage = mean(
      pourcentage_vehicules_lourds,
      na.rm = TRUE
    ),
    .groups = "drop"
  )

traffic |>
  select(
    annee,
    debit_total_24h,
    pourcentage_vehicules_lourds,
    estimated_heavy_vehicles
  ) |>
  cor(use = "complete.obs") |>
  round(2)

traffic |>
  ggplot(
    aes(
      x = debit_total_24h,
      y = pourcentage_vehicules_lourds,
      color = secteur
    )
  ) +
  geom_point(alpha = 0.35) +
  labs(
    title = "Total traffic and heavy-vehicle share",
    x = "Total 24-hour traffic volume",
    y = "Heavy vehicles (%)",
    color = "Sector"
  )
```

A cautious conclusion may compare the groups represented in the file. It should note that observed locations and years differ, and that the estimated heavy-vehicle count is calculated directly from total volume and percentage.
:::
