---
title: "Module 5 - Exercises"
subtitle: "STT-1100 Introduction to Data Science"
format:
  html:
    toc: true
    toc-title: Plan
    embed-resources: true
    link-external-newwindow: true
    css: [../css/base_css.css]
editor: visual
---

# Consolidation Exercises

These exercises are independent from the adventure and challenge. They consolidate the technical moves from module 5: manipulate dates, summarize groups, visualize associations, calculate simple correlations and write cautious conclusions.

The data used here are fictitious and do not represent any real system.

```{r setup, warning=FALSE, message=FALSE}
library(tidyverse)
library(lubridate)
```

# Readings to Review

- [R for Data Science - Exploratory data analysis](https://r4ds.hadley.nz/EDA.html){target="_blank"}
- [R for Data Science - Dates and times](https://r4ds.hadley.nz/datetimes.html){target="_blank"}
- [R for Data Science - Data visualization](https://r4ds.hadley.nz/data-visualize.html){target="_blank"}
- [R for Data Science - Missing values](https://r4ds.hadley.nz/missing-values.html){target="_blank"}
- [OpenIntro Statistics - Descriptive exploration](https://www.openintro.org/book/os/){target="_blank"}

After the readings, also complete the [formative mini-test](mini_test.qmd). It is not graded.

# Block A - Dates, Structure and Missing Values

The file `data/fictitious_support_workshops.csv` describes fake university support workshops.

## Exercise 1 - Import and Inspect

Import the file, display its structure and check its dimensions.

::: {.callout-note collapse="true"}
## Solution

```{r}
workshops <- read_csv(
  "data/fictitious_support_workshops.csv",
  show_col_types = FALSE
)

glimpse(workshops)

tibble(
  rows = nrow(workshops),
  columns = ncol(workshops)
)
```
:::

## Exercise 2 - Prepare Temporal Variables

Transform `workshop_date` into a date, then create `weekday`, `month` and `time_of_day`.

::: {.callout-note collapse="true"}
## Solution

```{r}
workshops_dates <- workshops |>
  mutate(
    workshop_date = ymd(workshop_date),
    weekday = wday(workshop_date, label = TRUE, abbr = FALSE),
    month = month(workshop_date, label = TRUE, abbr = FALSE),
    time_of_day = case_when(
      start_hour < 12 ~ "morning",
      start_hour < 17 ~ "afternoon",
      TRUE ~ "evening"
    )
  )

workshops_dates |>
  select(workshop_id, workshop_date, weekday, month, start_hour, time_of_day) |>
  slice_head(n = 8)
```
:::

## Exercise 3 - Identify Missing Values

Calculate the number of missing values in `participants`, `rain_mm` and `mean_satisfaction`.

::: {.callout-note collapse="true"}
## Solution

```{r}
workshops_dates |>
  summarise(
    participants_missing = sum(is.na(participants)),
    rain_missing = sum(is.na(rain_mm)),
    satisfaction_missing = sum(is.na(mean_satisfaction))
  )
```
:::

## Exercise 4 - Summarize by Time of Day

Calculate the number of workshops, mean participation and mean satisfaction by `time_of_day`.

::: {.callout-note collapse="true"}
## Solution

```{r}
summary_time <- workshops_dates |>
  group_by(time_of_day) |>
  summarise(
    n_workshops = n(),
    mean_participants = mean(participants, na.rm = TRUE),
    mean_satisfaction = mean(mean_satisfaction, na.rm = TRUE),
    .groups = "drop"
  )

summary_time
```
:::

# Block B - Associations and Visualizations

## Exercise 5 - Visualize Participants

Produce a column chart of the mean number of participants by time of day.

::: {.callout-note collapse="true"}
## Solution

```{r}
summary_time |>
  ggplot(aes(x = time_of_day, y = mean_participants)) +
  geom_col(fill = "#2c7fb8") +
  labs(
    title = "Mean participation by time of day",
    x = "Time of day",
    y = "Mean number of participants"
  )
```
:::

## Exercise 6 - Compare Topics

For each workshop topic, calculate the number of workshops, mean participation and mean satisfaction.

::: {.callout-note collapse="true"}
## Solution

```{r}
summary_topic <- workshops_dates |>
  group_by(topic) |>
  summarise(
    n_workshops = n(),
    mean_participants = mean(participants, na.rm = TRUE),
    mean_satisfaction = mean(mean_satisfaction, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(desc(mean_participants))

summary_topic
```
:::

## Exercise 7 - Calculate a Correlation Matrix

Calculate the correlations between `duration_minutes`, `participants`, `temperature_c`, `rain_mm` and `mean_satisfaction`.

::: {.callout-note collapse="true"}
## Solution

```{r}
workshops_dates |>
  select(duration_minutes, participants, temperature_c, rain_mm, mean_satisfaction) |>
  cor(use = "complete.obs") |>
  round(2)
```

Correlation summarizes a linear association. It is not enough to establish cause and effect.
:::

## Exercise 8 - Make a Scatterplot

Create a graph of `duration_minutes` and `participants`. Add a trend line.

::: {.callout-note collapse="true"}
## Solution

```{r}
workshops_dates |>
  ggplot(aes(x = duration_minutes, y = participants)) +
  geom_point(aes(color = topic), size = 2.5, alpha = 0.8) +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  labs(
    title = "Workshop duration and participation",
    x = "Workshop duration (minutes)",
    y = "Number of participants",
    color = "Topic"
  )
```
:::

## Exercise 9 - Compare Rainy and Non-Rainy Days

Create a `rainy_day` variable, then compare mean participation by that variable.

::: {.callout-note collapse="true"}
## Solution

```{r}
workshops_rain <- workshops_dates |>
  mutate(
    rainy_day = case_when(
      is.na(rain_mm) ~ "unknown rain",
      rain_mm > 0 ~ "rain",
      TRUE ~ "no rain"
    )
  )

workshops_rain |>
  group_by(rainy_day) |>
  summarise(
    n_workshops = n(),
    mean_participants = mean(participants, na.rm = TRUE),
    .groups = "drop"
  )
```
:::

# Block C - Cautious Interpretation

## Exercise 10 - Write a Descriptive Conclusion

Using one of the previous tables, write two sentences:

1. one sentence that describes the observed result;
2. one sentence that states a limitation.

::: {.callout-note collapse="true"}
## Example Answer

Evening workshops seem to have higher mean participation in this small fictitious file. This comparison remains descriptive because evening workshops do not always cover the same topics and the number of observations is limited.
:::

## Exercise 11 - Choose an Appropriate Visualization

Choose one of the two questions below, then produce an appropriate graph.

1. Does mean satisfaction vary by topic?
2. Does participation vary with temperature?

::: {.callout-note collapse="true"}
## Possible Solution

```{r}
workshops_dates |>
  filter(!is.na(mean_satisfaction)) |>
  ggplot(aes(x = topic, y = mean_satisfaction)) +
  geom_point(size = 2.5, alpha = 0.8) +
  labs(
    title = "Mean satisfaction by topic",
    x = "Topic",
    y = "Mean satisfaction"
  )
```
:::

## Exercise 12 - Build a Small Summary Table

Create a summary table by campus with the number of workshops, total participation, mean participation and mean satisfaction.

::: {.callout-note collapse="true"}
## Solution

```{r}
workshops_dates |>
  group_by(campus) |>
  summarise(
    n_workshops = n(),
    total_participation = sum(participants, na.rm = TRUE),
    mean_participation = mean(participants, na.rm = TRUE),
    mean_satisfaction = mean(mean_satisfaction, na.rm = TRUE),
    .groups = "drop"
  )
```
:::

# Case Studies

## Case Study 1 - Fictitious Library Visits

The file `data/fictitious_library_visits.csv` describes fake library-area visit observations.

Complete the following tasks:

1. import the file;
2. transform `visit_date` into a date;
3. create `weekday` and `time_of_day`;
4. summarize visitors by area and time of day;
5. calculate correlations between `visitors`, `temperature_c` and `rain_mm`;
6. produce a useful visualization;
7. write a cautious conclusion.

::: {.callout-note collapse="true"}
## Solution

```{r}
library_visits <- read_csv(
  "data/fictitious_library_visits.csv",
  show_col_types = FALSE
) |>
  mutate(
    visit_date = ymd(visit_date),
    weekday = wday(visit_date, label = TRUE, abbr = FALSE),
    time_of_day = case_when(
      hour < 12 ~ "morning",
      hour < 17 ~ "afternoon",
      TRUE ~ "evening"
    )
  )

library_visits |>
  group_by(area, time_of_day) |>
  summarise(
    n_observations = n(),
    mean_visitors = mean(visitors),
    .groups = "drop"
  )

library_visits |>
  select(visitors, temperature_c, rain_mm) |>
  cor(use = "complete.obs") |>
  round(2)

library_visits |>
  ggplot(aes(x = time_of_day, y = visitors, color = area)) +
  geom_point(size = 2.5, alpha = 0.8) +
  labs(
    title = "Fictitious visits by time of day",
    x = "Time of day",
    y = "Number of visitors",
    color = "Area"
  )
```

A cautious conclusion could state that evening observations seem busier in this file, especially in the computer area. This conclusion remains descriptive because exam week, area and campus vary at the same time.
:::

## Case Study 2 - Fictitious Campus Service Requests

The file `data/fictitious_campus_service_requests.csv` describes fake requests sent to campus services.

Complete the following tasks:

1. import the file;
2. transform `request_date` into a date;
3. summarize delays by service and priority;
4. calculate correlations between `delay_hours`, `satisfaction` and `estimated_cost`;
5. produce a useful visualization;
6. write a cautious conclusion.

::: {.callout-note collapse="true"}
## Solution

```{r}
requests <- read_csv(
  "data/fictitious_campus_service_requests.csv",
  show_col_types = FALSE
) |>
  mutate(
    request_date = ymd(request_date),
    weekday = wday(request_date, label = TRUE, abbr = FALSE)
  )

requests |>
  group_by(service, priority) |>
  summarise(
    n_requests = n(),
    mean_delay = mean(delay_hours, na.rm = TRUE),
    mean_satisfaction = mean(satisfaction, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(desc(mean_delay))

requests |>
  select(delay_hours, satisfaction, estimated_cost) |>
  cor(use = "complete.obs") |>
  round(2)

requests |>
  ggplot(aes(x = delay_hours, y = satisfaction, color = service)) +
  geom_point(size = 2.5, alpha = 0.8) +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  labs(
    title = "Processing delay and satisfaction",
    x = "Processing delay (hours)",
    y = "Satisfaction",
    color = "Service"
  )
```

A cautious conclusion could state that requests with longer delays seem associated with lower satisfaction in this fictitious file. However, we cannot conclude that delay directly causes satisfaction because priority, service type and request complexity may also matter.
:::
