---
title: "Consolidation exercises"
format:
  html:
    toc: true
    embed-resources: true
---

These exercises stabilize the core moves of the module: recognize a categorical variable, count levels, clean character strings, extract patterns and build readable charts.

They are independent from the adventure and challenge. They practise the same skills without directly preparing the food offence article.

```{r setup, warning=FALSE, message=FALSE}
library(tidyverse)
```

## Block A - Categorical variables and frequencies

### Exercise 1 - Recognize variable types

Here is a small dataset.

```{r}
survey <- tibble(
  program = c("Statistics", "Economics", "Statistics", "Biology", "Economics"),
  age = c(21, 24, 22, 20, 27),
  campus = c("Quebec City", "Quebec City", "Montreal", "Quebec City", "Montreal"),
  satisfaction = c("High", "Medium", "High", "Low", "Medium")
)
```

Identify which variables are categorical and which variables are numeric.

::: {.callout-note collapse="true"}
### Solution

`program`, `campus` and `satisfaction` are categorical. `age` is numeric.
:::

### Exercise 2 - Create a frequency table

Calculate the number of students by program.

::: {.callout-tip collapse="true"}
### Hint

Use `count()`.
:::

::: {.callout-note collapse="true"}
### Solution

```{r}
survey |>
  count(program, sort = TRUE)
```
:::

### Exercise 3 - Add proportions

Using the previous result, add a proportion and a percentage.

::: {.callout-note collapse="true"}
### Solution

```{r}
survey |>
  count(program, sort = TRUE) |>
  mutate(
    proportion = n / sum(n),
    percentage = 100 * proportion
  )
```
:::

### Exercise 4 - Build a cross-tabulation

Build a cross-tabulation between `program` and `campus`.

::: {.callout-note collapse="true"}
### Solution

```{r}
survey |>
  count(program, campus) |>
  pivot_wider(
    names_from = campus,
    values_from = n,
    values_fill = 0
  )
```
:::

## Block B - Character strings and patterns

### Exercise 5 - Standardize responses

Here is a response variable entered with inconsistent formats.

```{r}
responses <- tibble(
  response = c(" Yes", "yes", "YES ", "No", " no", "NO", "I do not know")
)
```

Create a variable `response_clean` in lowercase, without extra spaces.

::: {.callout-note collapse="true"}
### Solution

```{r}
responses |>
  mutate(response_clean = str_squish(str_to_lower(response)))
```
:::

### Exercise 6 - Detect a keyword

In the following vector, detect texts that contain `"delay"`, regardless of case.

```{r}
comments <- tibble(
  text = c(
    "Service delay",
    "Fast service",
    "major delay",
    "No issue",
    "REPEATED DELAY"
  )
)
```

::: {.callout-note collapse="true"}
### Solution

```{r}
comments |>
  mutate(contains_delay = str_detect(str_to_lower(text), "delay"))
```
:::

### Exercise 7 - Extract postal codes

From the following addresses, extract a simplified Canadian postal code.

```{r}
addresses <- tibble(
  address = c(
    "2500 boul. Laurier, Quebec City G1V 2L1",
    "475 Sherbrooke Street West, Montreal H3A 2L9",
    "100 Wellington Street, Gatineau J8X 2J1",
    "Unknown address"
  )
)
```

::: {.callout-note collapse="true"}
### Solution

```{r}
addresses |>
  mutate(
    postal_code = str_extract(
      address,
      "[A-Z][0-9][A-Z]\\s*[0-9][A-Z][0-9]"
    )
  )
```
:::

### Exercise 8 - Convert a text amount

Convert the following amounts to numeric values.

```{r}
fines <- tibble(
  amount = c("1 000 $", "500$", "2 250 $", "750 dollars")
)
```

::: {.callout-note collapse="true"}
### Solution

```{r}
fines |>
  mutate(fine_num = parse_number(
    amount,
    locale = locale(grouping_mark = " ")
  ))
```
:::

## Block C - Practise on fictitious campus service requests

Exercises 9 to 12 use a small fictitious dataset of campus service requests. This context is distinct from the adventure and challenge.

### Exercise 9 - Build a clean dataset

Here are the raw data.

```{r}
campus_services_raw <- tibble(
  id = paste0("REQ-", str_pad(1:12, 2, pad = "0")),
  campus = c("Quebec City", "Quebec", "Montreal", "Montréal", "Levis", "Lévis",
             "Quebec City", "Montreal", "Levis", "Quebec", "Montréal", "Lévis"),
  service = c("Library", "library", "Wi-Fi", "wifi", "Parking", "parking",
              "Shuttle", "shuttle", "Printing", "printing", "Library", "Wi-Fi"),
  description = c(
    "Shuttle delay reported",
    "Noisy reading room",
    "wifi unstable in building",
    "WiFi connection unavailable",
    "Payment impossible",
    "parking lot full",
    "Evening shuttle delay",
    "Crowded shuttle",
    "Printer without paper",
    "Printing blocked",
    "books misplaced",
    "Slow connection"
  ),
  status = c("open", "Open ", "closed", "Closed", "open", "in progress",
             "In progress", "closed", "open", "Closed", "in progress", "open"),
  cost = c("0 $", "150 $", "0$", "300 $", "1 200 $", "0 $",
           "850 $", "0 $", "75 $", "90 $", "0 $", "200 $")
)

campus_services_raw
```

Create a table `campus_services` containing:

- a standardized campus;
- a standardized service;
- a standardized status;
- a numeric cost variable `cost_num`.

::: {.callout-note collapse="true"}
### Solution

```{r}
campus_services <- campus_services_raw |>
  mutate(
    campus = str_squish(campus),
    campus = recode(
      campus,
      "Quebec" = "Quebec City",
      "Montréal" = "Montreal",
      "Lévis" = "Levis"
    ),
    service = str_to_lower(str_squish(service)),
    service = case_when(
      str_detect(service, "library") ~ "library",
      str_detect(service, "wi-fi|wifi") ~ "wi-fi",
      str_detect(service, "parking") ~ "parking",
      str_detect(service, "shuttle") ~ "shuttle",
      str_detect(service, "printing") ~ "printing",
      TRUE ~ service
    ),
    status = str_to_lower(str_squish(status)),
    cost_num = parse_number(cost, locale = locale(grouping_mark = " "))
  )

campus_services |>
  select(id, campus, service, status, cost_num)
```
:::

### Exercise 10 - Count requested services

Calculate the number of requests by campus and service.

::: {.callout-note collapse="true"}
### Solution

```{r}
campus_services |>
  count(campus, service, sort = TRUE)
```
:::

### Exercise 11 - Compare statuses by campus

Calculate the proportion of requests in each status for each campus.

::: {.callout-note collapse="true"}
### Solution

```{r}
campus_services |>
  count(campus, status) |>
  group_by(campus) |>
  mutate(proportion = n / sum(n)) |>
  ungroup() |>
  arrange(campus, desc(proportion))
```
:::

### Exercise 12 - Detect keywords

Create two logical variables:

- `contains_delay`, equal to `TRUE` if the description contains `"delay"`;
- `contains_connection`, equal to `TRUE` if the description contains `"wifi"` or `"connection"`.

::: {.callout-note collapse="true"}
### Solution

```{r}
campus_services <- campus_services |>
  mutate(
    description_lower = str_to_lower(description),
    contains_delay = str_detect(description_lower, "delay"),
    contains_connection = str_detect(description_lower, "wifi|connection")
  )

campus_services |>
  select(id, service, contains_delay, contains_connection)
```
:::

## Block D - Visualize categories

### Exercise 13 - Order a bar chart

Produce a bar chart showing the number of requests by service. Order services from most frequent to least frequent.

::: {.callout-note collapse="true"}
### Solution

```{r}
campus_services |>
  count(service, sort = TRUE) |>
  ggplot(aes(x = fct_reorder(service, n), y = n)) +
  geom_col(fill = "#2F6C8F") +
  coord_flip() +
  labs(
    x = "Service",
    y = "Number of requests",
    title = "Fictitious service requests by category"
  ) +
  theme_minimal()
```
:::

### Exercise 14 - Compare proportions with `position = "fill"`

Compare status composition between campuses.

::: {.callout-note collapse="true"}
### Solution

```{r}
campus_services |>
  count(campus, status) |>
  group_by(campus) |>
  mutate(proportion = n / sum(n)) |>
  ungroup() |>
  ggplot(aes(x = campus, y = proportion, fill = status)) +
  geom_col(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  labs(
    x = "Campus",
    y = "Proportion",
    fill = "Status",
    title = "Status composition by campus"
  ) +
  theme_minimal()
```
:::

### Exercise 15 - Compare costs by service

Create a chart comparing reported costs by service.

::: {.callout-note collapse="true"}
### Solution

```{r}
campus_services |>
  filter(cost_num > 0) |>
  ggplot(aes(x = service, y = cost_num)) +
  geom_boxplot(fill = "#D8A24A") +
  coord_flip() +
  scale_y_continuous(labels = scales::dollar) +
  labs(
    x = "Service",
    y = "Reported cost",
    title = "Costs associated with fictitious service requests"
  ) +
  theme_minimal()
```
:::

### Exercise 16 - Improve the message of a chart

Take one previous chart and improve it for a non-specialist reader:

- informative title;
- readable axes;
- ordered categories;
- units or percentages clearly shown;
- two-sentence comment under the chart.

::: {.callout-note collapse="true"}
### Example solution

```{r}
service_chart <- campus_services |>
  count(service, sort = TRUE) |>
  mutate(share = n / sum(n)) |>
  ggplot(aes(x = fct_reorder(service, share), y = share)) +
  geom_col(fill = "#2F6C8F") +
  coord_flip() +
  scale_y_continuous(labels = scales::percent) +
  labs(
    x = NULL,
    y = "Share of requests",
    title = "Requests are distributed across a few recurring services"
  ) +
  theme_minimal()

service_chart
```

Possible comment: the chart shows which services appear most often in this fictitious sample. However, the file is for method practice and does not measure a real volume of requests.
:::

## Case studies

### Case study 1 - Montréal 311 requests

The file `data/requetes_311_montreal.csv` is a real extract of Montréal 311 requests. A request does not necessarily represent a municipal intervention.

Complete the following tasks:

1. import the file;
2. clean the variable `categorie`;
3. calculate the number of requests by borough and category;
4. calculate status proportions by request nature;
5. produce a chart of the most frequent categories;
6. write two sentences that could appear in a short report.

::: {.callout-note collapse="true"}
### Solution

```{r}
requests_311 <- read_csv(
  "data/requetes_311_montreal.csv",
  show_col_types = FALSE
) |>
  mutate(categorie = str_squish(categorie))

requests_311 |>
  count(arrondissement, categorie, sort = TRUE)

requests_311 |>
  count(nature, statut) |>
  group_by(nature) |>
  mutate(proportion = n / sum(n)) |>
  ungroup()

requests_311 |>
  count(categorie, sort = TRUE) |>
  slice_head(n = 12) |>
  ggplot(aes(x = fct_reorder(categorie, n), y = n)) +
  geom_col(fill = "#4B8B3B") +
  coord_flip() +
  labs(
    x = "Category",
    y = "Number of requests",
    title = "Frequent categories in the 311 extract"
  ) +
  theme_minimal()
```

Possible comment: the extract describes published requests, not the real workload or quality of municipal interventions.
:::

### Case study 2 - Consumer complaints in Québec

The file `data/plaintes_consommation_quebec.csv` is a real extract of complaints received by Québec's Office de la protection du consommateur. A complaint is an unverified allegation.

Complete the following tasks:

1. import the file;
2. calculate the total number of complaints by head-office city;
3. identify the most frequent complaint reasons;
4. compare product categories between cities with a proportional chart;
5. propose an operational recommendation supported by a number.

::: {.callout-note collapse="true"}
### Solution

```{r}
transit_complaints <- read_csv(
  "data/plaintes_consommation_quebec.csv",
  show_col_types = FALSE
)

transit_complaints |>
  count(ville_siege_social, sort = TRUE)

transit_complaints |>
  count(motif, sort = TRUE)

transit_complaints |>
  count(ville_siege_social, categorie_produit, name = "total") |>
  group_by(ville_siege_social) |>
  mutate(proportion = total / sum(total)) |>
  ungroup() |>
  ggplot(aes(x = ville_siege_social, y = proportion, fill = categorie_produit)) +
  geom_col(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  labs(
    x = "Head-office city",
    y = "Proportion of complaints",
    fill = "Category",
    title = "Composition of consumer complaints"
  ) +
  theme_minimal()
```

Possible recommendation: describe the most frequent categories while explicitly noting that a complaint does not establish an offence.
:::
