---
title: "Consolidation Exercises"
format:
  html:
    toc: true
    embed-resources: true
---

# Consolidation exercises

```{r warning=FALSE, message=FALSE}
# Load the necessary libraries
library(tidyverse)
library(stringr)
```

----------------------------------------------------------------------------------

### Exercise 1 – Identify categorical variables

Here is a small dataset:

```{r}
data <- tibble::tibble(
  sex = c("Man", "Woman", "Woman", "Man", "Man"),
  age = c(25, 32, 29, 40, 23),
  region = c("East", "West", "East", "North", "South")
)
```

>  Among the three variables above, which are categorical?

::: {.callout-note collapse="true"}
### Reply

The `sex` and `region` variables are **categorical**. `age` is **numeric**.
:::

----------------------------------------------------------------------------------

### Exercise 2 – Frequency table

Creates a frequency table for the variable `sex`.

::: {.callout-tip collapse="true"}
### Track

Uses `count()` from the `dplyr` package.
:::

::: {.callout-note collapse="true"}
### Solution

```{r}
data %>% count(sex)
```
:::

----------------------------------------------------------------------------------

### Exercise 3 – Proportions

Calculates the percentages for each category of the `region` variable.

::: {.callout-tip collapse="true"}
### Track

Adds a column with `mutate(n / sum(n))`.
:::

::: {.callout-note collapse="true"}
### Solution

```{r}
data %>% count(region) %>% mutate(percent = n / sum(n) * 100)
```
:::

----------------------------------------------------------------------------------

### Exercise 4 – Barplot of a categorical variable

Make a barplot representing the `region` variable.

::: {.callout-tip collapse="true"}
### Track

Use `geom_bar()` with `aes(x = region)`.
:::

::: {.callout-note collapse="true"}
### Solution

```{r}
ggplot(data, aes(x = region)) + geom_bar()
```
:::

----------------------------------------------------------------------------------

### Exercise 5 – Comparison of two categorical variables

Adds a color by `sex` in the previous graph.

::: {.callout-note collapse="true"}
### Solution

```{r}
ggplot(data, aes(x = region, fill = sex)) + geom_bar(position = "dodge")
```
:::

----------------------------------------------------------------------------------

### Exercise 6 – Cleaning a categorical variable

Here is an inconsistent categorical variable:

```{r}
answers <- c("Yes", "yes", "No", "no", "YES", "NO")
```

Cleans up this variable so that all responses are lowercase **and without duplicates**.

::: {.callout-note collapse="true"}
### Solution

```{r}
unique(str_to_lower(answers))
```
:::

----------------------------------------------------------------------------------

### Exercise 7 – Crosstab

Here is another dataset:

```{r}
data2 <- tibble(
  smoker = c("Yes", "No", "No", "Yes", "Yes", "No", "No"),
  sport = c("Yes", "Yes", "No", "No", "Yes", "No", "No")
)
```

Creates a crosstab (contingency) between the two variables.

::: {.callout-note collapse="true"}
### Solution

```{r}
data2 %>% count(smoker, sport) %>% pivot_wider(names_from = sport, values_from = n, values_fill = 0)
```
:::

----------------------------------------------------------------------------------

### Exercise 8 – To go further: Mosaic plot

Uses the `ggplot2` package (or `ggmosaic` if available) to visually represent the relationship between `smoking` and `sport`.

::: {.callout-note collapse="true"}
### Solution with ggplot2

```{r}
# By grouping into grouped bars for an approximation
ggplot(data2, aes(x = smoker, fill = sport)) +
  geom_bar(position = "fill") +
  ylab("Proportion")
```
:::

----------------------------------------------------------------------------------

## Problems

### Exercise 9 – Palmer Penguins

We use the `penguins` dataset from the `palmerpenguins` package.

- Categorical variable: `species`, `island`, `sex`
- Numeric variable: `flipper_length_mm`, `body_mass_g`

Questions: 1. Create a frequency table for `species` 2. Visualize the distribution of species by island 3. Compare the sexes for each species

::: {.callout-note collapse="true"}
### Solution

```{r}
library(palmerpenguins)

penguins %>% count(species)

ggplot(penguins, aes(x = island, fill = species)) + geom_bar(position = "dodge")

ggplot(penguins, aes(x = species, fill = sex)) + geom_bar(position = "fill")
```
:::

----------------------------------------------------------------------------------

### Exercise 10 – Titanic Data

Use `titanic::titanic_train`

Categorical variables: `Sex`, `Embarked`, `Survived`

Questions: 1. Cross-tabulation between `Sex` and `Survived` 2. Proportion of survival by port of embarkation (`Embarked`) 3. Cross-tabulation `Sex` × `Embarked` × `Survived`

::: {.callout-note collapse="true"}
### Solution

```{r}
library(titanic)
titanic_train <- titanic::titanic_train

titanic_train %>% count(Sex, Survived)

titanic_train %>% count(Embarked, Survived) %>%
  group_by(Embarked) %>% mutate(p = n / sum(n))

ggplot(titanic_train, aes(x = Embarked, fill = factor(Survived))) +
  geom_bar(position = "fill") +
  facet_wrap(~Sex)
```
:::
