---
title: "STT-1100 — Checklist sheet"
author: "Laval University"
format:
  pdf:
    pagetemplate: cheatsheet
    keep-tex: true
    geometry: "landscape"
fontsize: 8pt
mainfont: "Arial"
logo: "../../css/STT-1100_hex.png"
---

```{r}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE, message = FALSE, warning = FALSE)
```


# WORK ENVIRONMENT

## RStudio
- Script: `Ctrl + Shift + N`
- Render Quarto: `Ctrl + Shift + K`

## Quarto
```markdown
---
title: "Title"
format: html
---
```

Chunk R:
````r
```{r}
summary(df)
```
````

## Basic packages
```r
library(tidyverse)
library(readr)
library(UlavalSSD)
library(palmerpenguins)
```

---

# MANIPULATE DATA — `dplyr`

## Basic

```r
select(df, col1)
filter(df, col1 > 10)
```

## Transformation

```r
mutate(df, z = x/y)
arrange(df, desc(x))
```

## Group and summarize

```r
group_by(df, cat)
summarise(df, avg = mean(x, na.rm = TRUE))
```

---

# VISUALIZATION — `ggplot2`

## Digital

```r
ggplot(df, aes(x, y)) + geom_point()
ggplot(df, aes(x)) + geom_histogram()
```

## Categorical

```r
ggplot(df, aes(x, fill = cat)) + geom_bar()
ggplot(df, aes(cat, num)) + geom_boxplot()
```

 Use `facet_wrap()` to compare multiple groups

---

# DESCRIPTIVE ANALYSIS

## Digital statistics

```r
mean(x, na.rm=TRUE)
sd(x)
summary(df)
```

## Frequencies

```r
table(df$cat)
prop.table(table(df$cat))
```

## `skimr`

```r
library(skimr)
skim(df)
```

---

# GOOD PRACTICE TIPS

- `glimpse(df)`: structure preview
- `head(df)` / `tail(df)`
- Comments `#` to clarify
- Frequent commits in GitHub
- Avoid ambiguous variable names
