---
title: "Consolidation exercises"
format:
  html:
    toc: true
    embed-resources: true
---

These exercises are independent from the adventure and challenge. They consolidate the technical moves of module 4: importing files, checking types, cleaning values, manipulating factors, reading JSON and documenting cleaning decisions.

```{r setup, warning=FALSE, message=FALSE}
library(readr)
library(dplyr)
library(tidyr)
library(stringr)
library(forcats)
library(tibble)
library(readxl)
library(jsonlite)
library(janitor)
```

## Block A - Import and inspect

### Exercise 1 - Read a CSV with controlled types

Import `policies.csv`, forcing `policy_id` to text and `premium_amount` to numeric.

::: {.callout-note collapse="true"}
### Solution

```{r}
policies <- read_csv(
  "policies.csv",
  col_types = cols(
    policy_id = col_character(),
    premium_amount = col_double(),
    .default = col_guess()
  )
)

glimpse(policies)
```
:::

### Exercise 2 - Check dimensions and names

Check the number of rows, number of columns and variable names in `policies`.

::: {.callout-note collapse="true"}
### Solution

```{r}
tibble(
  rows = nrow(policies),
  columns = ncol(policies)
)

names(policies)
```
:::

### Exercise 3 - Clean column names

Create a small table with inconvenient column names, then clean them with `clean_names()`.

::: {.callout-note collapse="true"}
### Solution

```{r}
raw_table <- tibble(
  "Program Name" = c("Statistics", "Computer Science"),
  "Requested Amount ($)" = c(2500, 1800)
)

clean_table <- raw_table |>
  clean_names()

names(clean_table)
```
:::

### Exercise 4 - Move from wide to long format

In `policies`, columns `q1_claims` to `q4_claims` represent quarterly claims. Transform them into two columns: `quarter` and `claims`.

::: {.callout-note collapse="true"}
### Solution

```{r}
policies_long <- policies |>
  pivot_longer(
    cols = q1_claims:q4_claims,
    names_to = "quarter",
    values_to = "claims"
  )

policies_long |>
  select(policy_id, quarter, claims) |>
  slice_head(n = 8)
```
:::

## Block B - Clean values and categories

### Exercise 5 - Return to wide format

Starting from `policies_long`, return to wide format.

::: {.callout-note collapse="true"}
### Solution

```{r}
policies_wide <- policies_long |>
  pivot_wider(
    names_from = quarter,
    values_from = claims
  )

glimpse(policies_wide)
```
:::

### Exercise 6 - Recode similar categories

The `vehicle_use` variable contains variants such as `"Pleasure"`, `"pleasure"` and `"Com.mute"`. Clean it to keep only `"Pleasure"` and `"Commute"`.

::: {.callout-note collapse="true"}
### Solution

```{r}
policies_use_clean <- policies |>
  mutate(
    vehicle_use = str_squish(vehicle_use),
    vehicle_use = fct_collapse(
      as_factor(vehicle_use),
      Pleasure = c("Pleasure", "pleasure"),
      Commute = c("Commute", "Com.mute")
    )
  )

policies_use_clean |>
  count(vehicle_use)
```
:::

### Exercise 7 - Parse an amount

Transform the following text amounts into numbers.

```{r}
amounts <- tibble(
  amount = c("1 234.56", "950.00", "-999", "2 500.25")
)
```

::: {.callout-note collapse="true"}
### Solution

```{r}
amounts |>
  mutate(
    amount_num = parse_number(
      amount,
      locale = locale(decimal_mark = ".", grouping_mark = " ")
    ),
    amount_num = na_if(amount_num, -999)
  )
```
:::

### Exercise 8 - Detect invalid FSA codes

An FSA code should contain three characters. In the following table, identify invalid rows.

```{r}
fsa_codes <- tibble(
  id = c("A", "B", "C", "D"),
  fsa_code = c("G1V", "H2X", "G1V1234", NA)
)
```

::: {.callout-note collapse="true"}
### Solution

```{r}
fsa_codes |>
  mutate(fsa_invalid = is.na(fsa_code) | nchar(fsa_code) != 3) |>
  filter(fsa_invalid)
```
:::

## Block C - Document cleaning

### Exercise 9 - Create a log structure

Create a list `cleaning_log` with categories `VM`, `VA`, `FT`, `RC` and `TY`.

::: {.callout-note collapse="true"}
### Solution

```{r}
cleaning_log <- list(
  VM = list(),
  VA = list(),
  FT = list(),
  RC = list(),
  TY = list()
)

str(cleaning_log)
```
:::

### Exercise 10 - Add a correction entry

Add an entry indicating that `-999` was replaced by `NA` in an amount variable.

::: {.callout-note collapse="true"}
### Solution

```{r}
cleaning_log$VM <- append(cleaning_log$VM, list(
  list(
    id = "exercise-7",
    variables = "amount",
    probleme = "Missing value coded as -999",
    action = "Replace with NA",
    justification = "The value -999 is not a real amount"
  )
))

cleaning_log$VM
```
:::

### Exercise 11 - Flag without correcting

Add an entry indicating that an invalid FSA code was flagged, but not corrected automatically.

::: {.callout-note collapse="true"}
### Solution

```{r}
cleaning_log$FT <- append(cleaning_log$FT, list(
  list(
    id = "C",
    variables = "fsa_code",
    probleme = "Invalid FSA code length",
    action = "Flag without automatic correction",
    justification = "The correct code cannot be inferred defensibly"
  )
))

cleaning_log$FT
```
:::

### Exercise 12 - Save a log to a temporary file

Save `cleaning_log` to a temporary file, then check that the file exists.

::: {.callout-note collapse="true"}
### Solution

```{r}
temp_path <- tempfile(fileext = ".Rdata")
save(cleaning_log, file = temp_path)
file.exists(temp_path)
```
:::

## Block D - Excel, JSON and validation rules

### Exercise 13 - Read a specific Excel sheet

The file `resources/quotes_2024.xlsx` contains a sheet named `Q3`. Import it while skipping the first row.

::: {.callout-note collapse="true"}
### Solution

```{r}
quotes_q3 <- read_excel(
  "resources/quotes_2024.xlsx",
  sheet = "Q3",
  skip = 1
)

glimpse(quotes_q3)
```
:::

### Exercise 14 - Read a simple JSON file

Read `coverage.json`, then retrieve the collision coverage limit.

::: {.callout-note collapse="true"}
### Solution

```{r}
coverage <- read_json("coverage.json")

coverage$coverage$collision$limit
```
:::

### Exercise 15 - Rectangle a JSON object

Transform the `coverage` section of the JSON file into a table.

::: {.callout-note collapse="true"}
### Solution

```{r}
coverage_tbl <- tibble(
  coverage_type = names(coverage$coverage),
  details = coverage$coverage
) |>
  unnest_wider(details)

coverage_tbl
```
:::

### Exercise 16 - Use a validation rule

The file `data/fictitious_equipment_rules.json` gives plausible installation years. Read this file and display the accepted bounds.

::: {.callout-note collapse="true"}
### Solution

```{r}
rules <- fromJSON("data/fictitious_equipment_rules.json")

tibble(
  year_min = rules$installation_year$min,
  year_max = rules$installation_year$max
)
```
:::

## Case studies

### Case study 1 - Fictitious scholarship applications

The file `data/fictitious_scholarship_applications.csv` contains fictitious student scholarship applications. It does not represent a real system.

Complete the following tasks:

1. import the file;
2. clean column names;
3. transform `requested_amount` into a number;
4. replace `annual_income = -999` with `NA`;
5. standardize program names;
6. identify invalid FSA codes;
7. create at least two cleaning-log entries.

::: {.callout-note collapse="true"}
### Solution

```{r}
scholarships <- read_csv(
  "data/fictitious_scholarship_applications.csv",
  show_col_types = FALSE
) |>
  clean_names() |>
  mutate(
    requested_amount_num = parse_number(
      requested_amount,
      locale = locale(decimal_mark = ".", grouping_mark = " ")
    ),
    annual_income = na_if(annual_income, -999),
    program_clean = str_to_lower(str_squish(program)),
    program_clean = case_when(
      str_detect(program_clean, "data|sci") ~ "data science",
      str_detect(program_clean, "math") ~ "mathematics",
      TRUE ~ program_clean
    ),
    status_clean = str_to_lower(str_squish(status)),
    fsa_invalid = is.na(fsa_code) | nchar(fsa_code) != 3
  )

scholarships |>
  count(program_clean, status_clean)

scholarships |>
  filter(fsa_invalid) |>
  select(application_id, fsa_code)

scholarship_log <- list(VM = list(), FT = list(), RC = list(), VA = list())

scholarship_log$VM <- append(scholarship_log$VM, list(
  list(
    id = "B-004",
    variables = "annual_income",
    probleme = "Missing value coded as -999",
    action = "Replace with NA",
    justification = "The value -999 is not a real annual income"
  )
))

scholarship_log$FT <- append(scholarship_log$FT, list(
  list(
    id = "B-010",
    variables = "fsa_code",
    probleme = "Invalid FSA code length",
    action = "Flag without automatic correction",
    justification = "The correct code cannot be inferred from the file"
  )
))

scholarship_log
```
:::

### Case study 2 - Fictitious municipal equipment records

The file `data/fictitious_municipal_equipment.csv` contains fictitious municipal equipment records. The file `data/fictitious_equipment_rules.json` contains validation rules.

Complete the following tasks:

1. import the CSV and JSON files;
2. clean `equipment_type` and `status`;
3. transform `maintenance_cost` into a number;
4. use the JSON rules to identify impossible years;
5. identify unexpected statuses, if any;
6. create a summary table by city;
7. document at least two decisions.

::: {.callout-note collapse="true"}
### Solution

```{r}
rules <- fromJSON("data/fictitious_equipment_rules.json")

equipment <- read_csv(
  "data/fictitious_municipal_equipment.csv",
  show_col_types = FALSE
) |>
  clean_names() |>
  mutate(
    equipment_type = str_to_lower(str_squish(equipment_type)),
    equipment_type = case_when(
      str_detect(equipment_type, "bus") ~ "bus shelter",
      str_detect(equipment_type, "play") ~ "play module",
      TRUE ~ equipment_type
    ),
    status = str_to_lower(str_squish(status)),
    maintenance_cost_num = parse_number(
      maintenance_cost,
      locale = locale(decimal_mark = ".", grouping_mark = " ")
    ),
    maintenance_cost_num = na_if(
      maintenance_cost_num,
      rules$maintenance_cost$coded_missing_value
    ),
    year_invalid = installation_year < rules$installation_year$min |
      installation_year > rules$installation_year$max,
    status_invalid = !status %in% rules$accepted_statuses,
    month_invalid = !inspection_month %in% rules$expected_inspection_months
  )

equipment |>
  filter(year_invalid | status_invalid | month_invalid) |>
  select(equipment_id, installation_year, status, inspection_month)

city_summary <- equipment |>
  group_by(city) |>
  summarise(
    n_equipment = n(),
    median_cost = median(maintenance_cost_num, na.rm = TRUE),
    n_invalid_years = sum(year_invalid),
    .groups = "drop"
  )

city_summary

equipment_log <- list(VM = list(), VA = list(), RC = list())

equipment_log$VA <- append(equipment_log$VA, list(
  list(
    id = equipment$equipment_id[equipment$year_invalid],
    variables = "installation_year",
    probleme = "Installation year outside JSON bounds",
    action = "Flag without automatic correction",
    justification = "The correct year cannot be inferred from the file"
  )
))

equipment_log$VM <- append(equipment_log$VM, list(
  list(
    id = "EQ-010",
    variables = "maintenance_cost",
    probleme = "Missing value coded as -999",
    action = "Replace with NA",
    justification = "The JSON rules state that -999 codes a missing value"
  )
))

equipment_log
```
:::
