---
title: "Consolidation exercises"
format:
  html:
    toc: true
    embed-resources: true
---

These exercises are independent from the adventure and challenge. They consolidate the technical moves of module 4: importing files, checking types, cleaning values, manipulating factors, reading JSON and documenting cleaning decisions.

::: {.callout-note}
## Two real Québec case studies

The case studies use an aggregated table on [last-resort financial assistance](https://www.donneesquebec.ca/recherche/dataset/aide-financiere-de-dernier-recours-afdr-clientele-et-prestations) published by Québec's Ministère de l'Emploi et de la Solidarité sociale and the [Sherbrooke sports and recreation facilities](https://www.donneesquebec.ca/recherche/dataset/b6498f3436974ecbb8fa636a7d9c0b2f_0) inventory. Both sources are released under the CC BY 4.0 licence. The local files are reproducible snapshots prepared by `scripts/build_module04_real_data.R`.
:::

```{r setup, warning=FALSE, message=FALSE}
library(readr)
library(dplyr)
library(tidyr)
library(stringr)
library(forcats)
library(tibble)
library(readxl)
library(jsonlite)
library(janitor)
```

## Block A - Import and inspect

### Exercise 1 - Read a CSV with controlled types

Import `policies.csv`, forcing `policy_id` to text and `premium_amount` to numeric.

::: {.callout-note collapse="true"}
### Solution

```{r}
policies <- read_csv(
  "policies.csv",
  col_types = cols(
    policy_id = col_character(),
    premium_amount = col_double(),
    .default = col_guess()
  )
)

glimpse(policies)
```
:::

### Exercise 2 - Check dimensions and names

Check the number of rows, number of columns and variable names in `policies`.

::: {.callout-note collapse="true"}
### Solution

```{r}
tibble(
  rows = nrow(policies),
  columns = ncol(policies)
)

names(policies)
```
:::

### Exercise 3 - Clean column names

Create a small table with inconvenient column names, then clean them with `clean_names()`.

::: {.callout-note collapse="true"}
### Solution

```{r}
raw_table <- tibble(
  "Program Name" = c("Statistics", "Computer Science"),
  "Requested Amount ($)" = c(2500, 1800)
)

clean_table <- raw_table |>
  clean_names()

names(clean_table)
```
:::

### Exercise 4 - Move from wide to long format

In `policies`, columns `q1_claims` to `q4_claims` represent quarterly claims. Transform them into two columns: `quarter` and `claims`.

::: {.callout-note collapse="true"}
### Solution

```{r}
policies_long <- policies |>
  pivot_longer(
    cols = q1_claims:q4_claims,
    names_to = "quarter",
    values_to = "claims"
  )

policies_long |>
  select(policy_id, quarter, claims) |>
  slice_head(n = 8)
```
:::

## Block B - Clean values and categories

### Exercise 5 - Return to wide format

Starting from `policies_long`, return to wide format.

::: {.callout-note collapse="true"}
### Solution

```{r}
policies_wide <- policies_long |>
  pivot_wider(
    names_from = quarter,
    values_from = claims
  )

glimpse(policies_wide)
```
:::

### Exercise 6 - Recode similar categories

The `vehicle_use` variable contains variants such as `"Pleasure"`, `"pleasure"` and `"Com.mute"`. Clean it to keep only `"Pleasure"` and `"Commute"`.

::: {.callout-note collapse="true"}
### Solution

```{r}
policies_use_clean <- policies |>
  mutate(
    vehicle_use = str_squish(vehicle_use),
    vehicle_use = fct_collapse(
      as_factor(vehicle_use),
      Pleasure = c("Pleasure", "pleasure"),
      Commute = c("Commute", "Com.mute")
    )
  )

policies_use_clean |>
  count(vehicle_use)
```
:::

### Exercise 7 - Parse an amount

Transform the following text amounts into numbers.

```{r}
amounts <- tibble(
  amount = c("1 234.56", "950.00", "-999", "2 500.25")
)
```

::: {.callout-note collapse="true"}
### Solution

```{r}
amounts |>
  mutate(
    amount_num = parse_number(
      amount,
      locale = locale(decimal_mark = ".", grouping_mark = " ")
    ),
    amount_num = na_if(amount_num, -999)
  )
```
:::

### Exercise 8 - Detect invalid FSA codes

An FSA code should contain three characters. In the following table, identify invalid rows.

```{r}
fsa_codes <- tibble(
  id = c("A", "B", "C", "D"),
  fsa_code = c("G1V", "H2X", "G1V1234", NA)
)
```

::: {.callout-note collapse="true"}
### Solution

```{r}
fsa_codes |>
  mutate(fsa_invalid = is.na(fsa_code) | nchar(fsa_code) != 3) |>
  filter(fsa_invalid)
```
:::

## Block C - Document cleaning

### Exercise 9 - Create a log structure

Create a list `cleaning_log` with categories `VM`, `VA`, `FT`, `RC` and `TY`.

::: {.callout-note collapse="true"}
### Solution

```{r}
cleaning_log <- list(
  VM = list(),
  VA = list(),
  FT = list(),
  RC = list(),
  TY = list()
)

str(cleaning_log)
```
:::

### Exercise 10 - Add a correction entry

Add an entry indicating that `-999` was replaced by `NA` in an amount variable.

::: {.callout-note collapse="true"}
### Solution

```{r}
cleaning_log$VM <- append(cleaning_log$VM, list(
  list(
    id = "exercise-7",
    variables = "amount",
    probleme = "Missing value coded as -999",
    action = "Replace with NA",
    justification = "The value -999 is not a real amount"
  )
))

cleaning_log$VM
```
:::

### Exercise 11 - Flag without correcting

Add an entry indicating that an invalid FSA code was flagged, but not corrected automatically.

::: {.callout-note collapse="true"}
### Solution

```{r}
cleaning_log$FT <- append(cleaning_log$FT, list(
  list(
    id = "C",
    variables = "fsa_code",
    probleme = "Invalid FSA code length",
    action = "Flag without automatic correction",
    justification = "The correct code cannot be inferred defensibly"
  )
))

cleaning_log$FT
```
:::

### Exercise 12 - Save a log to a temporary file

Save `cleaning_log` to a temporary file, then check that the file exists.

::: {.callout-note collapse="true"}
### Solution

```{r}
temp_path <- tempfile(fileext = ".Rdata")
save(cleaning_log, file = temp_path)
file.exists(temp_path)
```
:::

## Block D - Excel, JSON and validation rules

### Exercise 13 - Read a specific Excel sheet

The file `resources/quotes_2024.xlsx` contains a sheet named `Q3`. Import it while skipping the first row.

::: {.callout-note collapse="true"}
### Solution

```{r}
quotes_q3 <- read_excel(
  "resources/quotes_2024.xlsx",
  sheet = "Q3",
  skip = 1
)

glimpse(quotes_q3)
```
:::

### Exercise 14 - Read a simple JSON file

Read `coverage.json`, then retrieve the collision coverage limit.

::: {.callout-note collapse="true"}
### Solution

```{r}
coverage <- read_json("coverage.json")

coverage$coverage$collision$limit
```
:::

### Exercise 15 - Rectangle a JSON object

Transform the `coverage` section of the JSON file into a table.

::: {.callout-note collapse="true"}
### Solution

```{r}
coverage_tbl <- tibble(
  coverage_type = names(coverage$coverage),
  details = coverage$coverage
) |>
  unnest_wider(details)

coverage_tbl
```
:::

### Exercise 16 - Read official source metadata

The file `data/metadonnees_installations_sherbrooke.json` comes from the City of Sherbrooke REST service. Read it, then display the field names and the coordinate system used for the extent.

::: {.callout-note collapse="true"}
### Solution

```{r}
metadata <- fromJSON(
  "data/metadonnees_installations_sherbrooke.json"
)

metadata$fields |>
  select(name, type, alias)

metadata$extent$spatialReference$latestWkid
```
:::

## Case studies

### Case study 1 - Last-resort financial assistance in Québec

The file `data/afdr_clientele_prestations_2022_12.csv` contains 43 aggregated rows published by Québec's Ministère de l'Emploi et de la Solidarité sociale for December 2022. Each row describes a client characteristic or a region. It contains no individual records.

Complete the following tasks:

1. import every column as text;
2. clean the column names;
3. convert the numeric columns to the correct type;
4. convert `caracteristique` to a factor;
5. check that all numbers are non-negative;
6. check that `nb_prestataires` equals the sum of adults and children;
7. produce a table of regions ranked by total benefits paid;
8. document at least two decisions in a cleaning log.

::: {.callout-note collapse="true"}
### Solution

```{r}
afdr_raw <- read_csv(
  "data/afdr_clientele_prestations_2022_12.csv",
  col_types = cols(.default = col_character())
) |>
  clean_names()

afdr <- afdr_raw |>
  mutate(
    across(nb_menages:prest_base_heberges, parse_double),
    caracteristique = as_factor(caracteristique),
    valeur = str_squish(valeur),
    negative_value = if_any(
      nb_menages:prest_base_heberges,
      ~ .x < 0
    ),
    inconsistent_total = nb_prestataires != nb_adultes + nb_enfants
  )

afdr |>
  count(caracteristique)

afdr |>
  summarise(
    n_negative_values = sum(negative_value),
    n_inconsistent_totals = sum(inconsistent_total)
  )

afdr_regions <- afdr |>
  filter(caracteristique == "Région") |>
  select(valeur, nb_prestataires, prestation_totale_versee) |>
  arrange(desc(prestation_totale_versee))

afdr_regions

afdr_log <- list(TY = list(), RC = list(), VA = list())

afdr_log$TY <- append(afdr_log$TY, list(
  list(
    id = "afdr-file-2022-12",
    variables = "nb_menages:prest_base_heberges",
    probleme = "Columns deliberately imported as text",
    action = "Controlled conversion with parse_double()",
    justification = "Calculations require numeric columns"
  )
))

afdr_log$VA <- append(afdr_log$VA, list(
  list(
    id = "afdr-file-2022-12",
    variables = "nb_prestataires",
    probleme = "Total consistency must be checked",
    action = "Compare with nb_adultes + nb_enfants",
    justification = "The relationship can be tested without altering the source"
  )
))

afdr_log
```
:::

### Case study 2 - Sherbrooke sports and recreation facilities

The file `data/installations_sportives_sherbrooke.csv` contains the municipal inventory published by the City of Sherbrooke. The file `data/metadonnees_installations_sherbrooke.json` is the official description of the corresponding ArcGIS service. A missing value in `nom`, `detail` or `surface` is not automatically an error.

Complete the following tasks:

1. import the CSV and JSON files;
2. clean column names and extra whitespace;
3. convert `eclairage` to a factor;
4. check that the official JSON fields are present in the CSV;
5. use the JSON extent to identify out-of-range coordinates;
6. create a summary table by facility type;
7. flag missing names without inventing them;
8. document at least two decisions.

::: {.callout-note collapse="true"}
### Solution

```{r}
metadata <- fromJSON(
  "data/metadonnees_installations_sherbrooke.json"
)

facilities_raw <- read_csv(
  "data/installations_sportives_sherbrooke.csv",
  show_col_types = FALSE
)

expected_fields <- str_to_lower(metadata$fields$name)
missing_fields <- setdiff(
  expected_fields,
  names(clean_names(facilities_raw))
)

facilities <- facilities_raw |>
  clean_names() |>
  mutate(
    across(c(type, detail, nom, surface, eclairage), str_squish),
    eclairage = factor(eclairage, levels = c("Non", "Oui")),
    missing_name = is.na(nom),
    outside_extent = x < metadata$extent$xmin |
      x > metadata$extent$xmax |
      y < metadata$extent$ymin |
      y > metadata$extent$ymax
  )

missing_fields

facilities |>
  summarise(
    n_outside_extent = sum(outside_extent, na.rm = TRUE),
    n_missing_names = sum(missing_name)
  )

type_summary <- facilities |>
  group_by(type) |>
  summarise(
    n_facilities = n(),
    n_missing_names = sum(missing_name),
    illuminated_share = mean(eclairage == "Oui", na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(desc(n_facilities))

type_summary

facilities_log <- list(VM = list(), FT = list(), RC = list())

facilities_log$FT <- append(facilities_log$FT, list(
  list(
    id = facilities$objectid[facilities$missing_name],
    variables = "nom",
    probleme = "Name missing from the source",
    action = "Flag without automatic correction",
    justification = "A name cannot be inferred from the other fields"
  )
))

facilities_log$RC <- append(facilities_log$RC, list(
  list(
    id = "sherbrooke-facilities-file",
    variables = "eclairage",
    probleme = "Categorical variable imported as text",
    action = "Convert to a Non/Oui factor",
    justification = "Observed levels are explicit and preserved"
  )
))

facilities_log
```
:::
