---
title: "Adventure 9 - Algorithmic prediction and bias"
subtitle: "STT-1100 Introduction to Data Science"
author:
  - institute: "Laval University"
editor: visual
editor_options:
  chunk_output_type: console
format:
  html:
    toc: true
    toc-title: Plan
    css: [../css/base_css.css]
    self-contained: true
---

```{r setup, include=FALSE}
library(tidyverse)
```

# Scenario: Quality control in education

You work as a **junior data scientist** in the quality control and equity department at the Quebec Ministry of Education. Your mandate: use real and fictitious data to identify, evaluate and mitigate algorithmic bias in educational decisions.

Your superior, **Michel Dufour**, deputy director of the department, welcomes you with these words:

> "Welcome to the team!" Our work is crucial: ensuring that every child has equal opportunities to succeed. To do this, we not only need to effectively predict certain indicators, but also understand and avoid biases in our models. »

::: {.callout-note}
**Your contact: Michel Dufour**\
He will accompany you on this adventure: he will ask key questions and await your results and detailed analyses.
:::

## Adventure Objectives

- Build a clear predictive model from real data.
- Identify and explain potential biases on fictitious data.
- Clearly present your analyzes and thoughts on these biases.
- Propose concrete recommendations to mitigate these biases.

----------------------------------------------------------------------------------

# Mission 1: Predictive model at school level

Michel gives you a [data file from the Data Québec portal](https://www.donneesquebec.ca/recherche/dataset/indices-de-defavorisation/resource/6c5d4a5d-ba3b-40a6-b570-916f43ab622c). These data contain information on disadvantage on primary schools in Quebec.

::: {.callout-important title="Michel explains to you"}
"I invite you to explore the data. But be careful: the column names are not always intuitive. Start by visiting the documentation page on Data Québec to fully understand their meaning. »
:::

## Step 1: Exploration and cleaning

- Identify relevant variables.
- Do a summary cleanup if necessary (missing values, data types, etc.).
- Carry out a complete descriptive analysis of this data (refer to modules 2 to 4).

```{r data-load, eval=FALSE}
# df_ecoles <- read_csv("ecoles_primaires_qc.csv") # File missing
glimpse(df_schools)
```

## Step 2: Regression Model

::: {.callout-important title="Michel asks you"}
"I would like you to build a linear regression model to predict the **IMSE** (Socio-Economic Environment Index), based on two variables: the proportion of low-income students (**SFR**) and the **total number of students** in the school. »
:::

```{r regression, eval=FALSE}
model <- lm( ... , data = ...)
summary(model)
```

::: {.callout-important title="Michel questions you"}
"Once the model is built, can you explain to me, in simple language, what the two coefficients mean? »
:::

## Step 3: Targeted Predictions

You will then use the `predict()` function to generate IMSE predictions from your model.

```{r prediction-example, eval=FALSE}
predict(model, newdata = df_schools)
```

::: {.callout-important title="Michel specifies"}
"I would like you to use your template to:"

\- predict the IMSE for the **10 largest schools** (in number of students);

\- predict the IMSE for the **10 schools with the highest SFRs**.

Compare predicted values ​​to actual values. Is your model good?
:::

## Step 4: Missing School Predictions

::: {.callout-important title="Michel adds"}
"Last thing: in the data, some schools have the variable **Diffusion** set to "NO". These schools do not have a measured IMSE — often because they have fewer than 30 students. »

"Can you use your model to **predict the IMSE of all schools with fewer than 30 students**? Next, I would like to know: what do you think about the validity of these predictions? Can we trust it? »
:::

----------------------------------------------------------------------------------

# Mission 2: Detection of individual biases

Michel also provides you with a fictitious dataset of 1000 students. At first glance, this game seems neutral. But is this really the case?

Load the data:

```{r dummy-load, eval=FALSE}
df_students <- read_csv("students_fictifs.csv")
glimpse(df_students)
```

### Step 1: Overall analysis

Start by calculating and visualizing the **overall admission rate**.

```{r admission-global, eval=FALSE}
df_students %>%
  count(Admission) %>%
  mutate(pct = n / sum(n))
```

Discuss what you observe: is it balanced? Surprising?

### Step 2: Analysis by group

Now, explore admission rates **by gender**.

```{r admission-by-sex, eval=FALSE}
# code to modify
df_students %>%
  count(Admission) %>%
  group_by(Sex) %>%
  mutate(pct = n / sum(n))
```

Produce a clear graph to illustrate potential differences.

### Step 3: Guided discussion

- Do you see a notable difference between the groups?
- Is this difference justified? What assumptions can you make?
- Can we talk about **bias** here? For what ?

### Step 4: Exploring other variables

Then continue with the analysis according to other variables (mother tongue, socio-economic level, access to resources).

You can reproduce the same logic: crosstab + graph + interpretation.

**To be submitted:** a brief summary presenting the bias(es) detected, with clear visual support and recommendations or avenues for improvement.

----------------------------------------------------------------------------------

# Challenge to submit

You will have to produce a **video capsule of 180 seconds maximum** in which you present:

- Or your predictive model based on real data (Mission 1).
- Either your analysis of the bias detected from the fictitious data (Mission 2).

Your capsule must include:

- A short introduction (put into context).
- Your methodology briefly explained.
- Your results presented visually (tables, graphs).
- A clear conclusion with recommendations or avenues for improvement.

Your video capsule must be uploaded to the indicated GitHub repository.

::: {.callout-tip}
**Tips for making your video successful:**

- Be concise and clear.
- Rely on readable graphics.
- Speak slowly, clearly, and be dynamic!
:::

----------------------------------------------------------------------------------

# Conclusion of the adventure

You have built a predictive model useful for decision making and learned to detect critical biases in individual data.

Bravo for your rigor, your sense of observation and your ethical posture!
