---
title: "Consolidation Exercises"
format:
  html:
    toc: true
    embed-resources: true
execute:
  warning: false
  message: false
---

# Consolidation Exercises {.unnumbered}

These exercises are independent from the adventure and the challenge. They consolidate linear regression, prediction and careful discussion of bias using fictitious datasets that differ from the main activity.

Before starting, review the module resources if needed: chapters 7 and 8 of *Introduction to Modern Statistics*, the `lm()` documentation, the `predict.lm()` documentation and the module formative mini-test.

```{r setup}
library(tidyverse)
```

## Block A - Explore a linear relationship

### Exercise 1 - Import the data

Import the file `fictitious_housing_energy.csv`. Each row represents a fictitious housing unit in the Quebec City region.

```{r}
housing <- read_csv("data/fictitious_housing_energy.csv", show_col_types = FALSE)

glimpse(housing)
```

::: {.callout-tip collapse="true"}
The table contains 32 fictitious housing units. The variable `annual_kwh` will be used as the response variable in the first exercises.
:::

### Exercise 2 - Visualize the main relationship

Plot the relationship between housing area and annual electricity consumption. Add a regression line.

```{r}
ggplot(housing, aes(x = area_m2, y = annual_kwh)) +
  geom_point(alpha = 0.75, size = 2) +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE) +
  labs(
    x = "Housing area (m2)",
    y = "Annual consumption (kWh)",
    title = "Annual consumption by housing area"
  ) +
  theme_minimal()
```

::: {.callout-tip collapse="true"}
The relationship appears positive: larger housing units tend to have higher annual consumption. This chart does not prove that area causes all consumption, but it suggests an association that is useful for a first model.
:::

### Exercise 3 - Fit a simple regression

Fit a model that predicts annual consumption from area.

```{r}
simple_model <- lm(annual_kwh ~ area_m2, data = housing)

summary(simple_model)
```

Interpret the slope in your own words.

::: {.callout-tip collapse="true"}
The slope estimates the average predicted change in annual consumption when area increases by one square metre. The interpretation should remain descriptive because other variables such as occupants, building age and heating type may also play a role.
:::

## Block B - Predict and diagnose

### Exercise 4 - Fit a multiple regression

Now fit a model that adds the number of occupants and the year built.

```{r}
multiple_model <- lm(
  annual_kwh ~ area_m2 + occupants + year_built,
  data = housing
)

summary(multiple_model)
```

Compare the interpretation of the `area_m2` coefficient with the one obtained in the simple model.

::: {.callout-tip collapse="true"}
In the multiple model, the `area_m2` coefficient is interpreted while holding `occupants` and `year_built` fixed. It therefore does not answer exactly the same question as the simple model.
:::

### Exercise 5 - Produce predictions

Create two new fictitious housing units, then use the multiple model to predict their annual consumption.

```{r}
new_housing <- tibble(
  area_m2 = c(70, 120),
  occupants = c(2, 4),
  year_built = c(1990, 2015)
)

housing_predictions <- new_housing |>
  mutate(predicted_consumption = predict(multiple_model, newdata = new_housing))

housing_predictions
```

::: {.callout-tip collapse="true"}
Predictions are expected values according to the model. They do not replace a real measurement, and they become more fragile if the new housing unit looks very different from the units used to fit the model.
:::

### Exercise 6 - Compare observed and predicted values

Add the multiple-model predictions to the original table, then calculate the error `observed - predicted`.

```{r}
model_diagnostics <- housing |>
  mutate(
    predicted_consumption = predict(multiple_model, newdata = housing),
    error_kwh = annual_kwh - predicted_consumption,
    absolute_error_kwh = abs(error_kwh)
  )

model_diagnostics |>
  arrange(desc(absolute_error_kwh)) |>
  select(
    housing_id,
    neighbourhood,
    annual_kwh,
    predicted_consumption,
    error_kwh,
    absolute_error_kwh
  ) |>
  slice_head(n = 6)
```

Then calculate a root mean squared error.

```{r}
rmse_kwh <- sqrt(mean(model_diagnostics$error_kwh^2))

rmse_kwh
```

::: {.callout-tip collapse="true"}
Rows with the largest absolute errors deserve inspection. They may reveal unusual housing units, important variables missing from the model or simply natural variability.
:::

### Exercise 7 - Visualize errors

Plot the errors against the predicted values. The chart should help you see whether errors remain centred around zero.

```{r}
ggplot(model_diagnostics, aes(x = predicted_consumption, y = error_kwh)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_point(alpha = 0.75, size = 2) +
  labs(
    x = "Predicted consumption (kWh)",
    y = "Error: observed - predicted (kWh)",
    title = "Visual diagnosis of errors"
  ) +
  theme_minimal()
```

::: {.callout-tip collapse="true"}
A useful reflex is to look for patterns: errors that increase with predictions, separated groups or highly isolated points. Such a pattern does not automatically invalidate the model, but it signals a limit to discuss.
:::

## Block C - Examine descriptive bias

### Exercise 8 - Import a second dataset

Import the file `fictitious_culture_grants.csv`. Each row represents a fictitious cultural grant application.

```{r}
grants <- read_csv("data/fictitious_culture_grants.csv", show_col_types = FALSE)

glimpse(grants)
```

::: {.callout-tip collapse="true"}
This dataset is fictitious. It is used to practise descriptive reading of differences between groups without concluding too quickly about cause or discrimination.
:::

### Exercise 9 - Compare rates by region

Calculate the acceptance rate by region, then produce a chart.

```{r}
acceptance_by_region <- grants |>
  group_by(region) |>
  summarise(
    applications = n(),
    acceptance_rate = mean(decision == "Accepted"),
    .groups = "drop"
  ) |>
  arrange(acceptance_rate)

acceptance_by_region
```

```{r}
ggplot(acceptance_by_region, aes(x = acceptance_rate, y = reorder(region, acceptance_rate))) +
  geom_col() +
  labs(
    x = "Acceptance rate",
    y = NULL,
    title = "Acceptance rate by region"
  ) +
  theme_minimal()
```

::: {.callout-tip collapse="true"}
Raw rates can draw attention to differences, but the groups have few observations. Avoid a definitive conclusion.
:::

### Exercise 10 - Add context variables

Also compare decisions by organization size and by whether the application is a first application.

```{r}
acceptance_by_size <- grants |>
  group_by(organization_size) |>
  summarise(
    applications = n(),
    acceptance_rate = mean(decision == "Accepted"),
    mean_score = mean(committee_score),
    .groups = "drop"
  )

acceptance_by_size
```

```{r}
acceptance_by_first_application <- grants |>
  group_by(first_application) |>
  summarise(
    applications = n(),
    acceptance_rate = mean(decision == "Accepted"),
    median_budget = median(requested_budget),
    mean_score = mean(committee_score),
    .groups = "drop"
  )

acceptance_by_first_application
```

::: {.callout-tip collapse="true"}
A regional difference may be linked to other available variables, such as organization size, committee score or first-application status. This does not settle the bias question, but it makes the analysis more honest.
:::

### Exercise 11 - Write a careful conclusion

Using the previous tables, write three sentences:

1. one sentence that describes an observed difference;
2. one sentence that mentions a data limitation;
3. one sentence that proposes an additional check.

::: {.callout-tip collapse="true"}
Possible example: acceptance rates vary between regions in this fictitious file. However, each region contains few applications and raw rates do not account for all application characteristics. A next step would be to check whether differences remain among applications with comparable scores, comparable disciplines and comparable organization sizes.
:::

## Case Study 1 - Predict residential consumption

A small municipal organization wants to forecast the annual consumption of new housing units in order to plan an energy-support program.

Your task:

1. choose between `simple_model` and `multiple_model`;
2. predict consumption for three fictitious housing units;
3. write three limits to communicate before using the predictions.

```{r}
housing_to_evaluate <- tibble(
  area_m2 = c(65, 95, 145),
  occupants = c(1, 3, 5),
  year_built = c(1972, 2001, 2018)
)

housing_to_evaluate |>
  mutate(predicted_consumption = predict(multiple_model, newdata = housing_to_evaluate))
```

::: {.callout-tip collapse="true"}
The multiple model is preferable here because it uses more relevant information. Limits to communicate may include: the data are fictitious and small; the model does not include insulation, heating habits or appliances; predictions should remain within the range of housing units similar to those observed.
:::

## Case Study 2 - Audit a grant procedure

A committee wants to know whether its grant decisions show concerning differences between groups.

Your task:

1. choose two comparison variables;
2. calculate acceptance rates and mean scores;
3. write a short paragraph that distinguishes descriptive signal, limit and next analysis.

```{r}
audit_summary <- grants |>
  group_by(region, organization_size) |>
  summarise(
    applications = n(),
    acceptance_rate = mean(decision == "Accepted"),
    mean_score = mean(committee_score),
    median_budget = median(requested_budget),
    .groups = "drop"
  ) |>
  arrange(region, organization_size)

audit_summary
```

::: {.callout-tip collapse="true"}
A careful paragraph could say that acceptance rates differ across some combinations of region and organization size in the fictitious data. However, several cells contain few applications, which makes comparisons unstable. Before discussing bias or discrimination, comparable applications should be examined, decision criteria should be documented and possible proxy variables should be checked.
:::
