---
title: "Consolidation Exercises"
format:
  html:
    toc: true
    embed-resources: true
execute:
  warning: false
  message: false
---

These exercises are independent from the adventure and the challenge. They consolidate text analysis, lexical scores, TF-IDF and dashboard preparation using real public Quebec text.

The first file contains a balanced sample of descriptions from the [official Données Québec API](https://www.donneesquebec.ca/page-api/){target="_blank"}. The second comes from the Ministry of Tourism dataset [Événements - Système d'information touristique Québec](https://www.donneesquebec.ca/recherche/dataset/sit-quebec-evenements){target="_blank"}. Both sources are distributed under the CC BY 4.0 licence.

```{r setup}
library(tidyverse)
library(tidytext)
library(stringr)
library(stopwords)
library(forcats)
```

## Block A - Turn descriptions into tokens

### Exercise 1 - Import descriptions

Import `quebec_dataset_descriptions.csv`. Each row is a real portal record with its title, description, producer and category.

```{r}
catalog <- read_csv(
  "data/quebec_dataset_descriptions.csv",
  show_col_types = FALSE
)

glimpse(catalog)
```

::: {.callout-tip collapse="true"}
The sample contains 89 records from eight cities: Quebec City, Gatineau, Sherbrooke, Trois-Rivières, Saguenay, Laval, Longueuil and Montreal. Montreal accounts for one producer out of eight.
:::

### Exercise 2 - Check the structure

```{r}
catalog |>
  group_by(producer) |>
  summarise(
    records = n(),
    categories = n_distinct(category),
    .groups = "drop"
  ) |>
  arrange(desc(records), producer)
```

::: {.callout-tip collapse="true"}
The selection is balanced by producer, but does not represent the full catalog. It favours recently updated records with descriptions of at least 50 characters.
:::

### Exercise 3 - Tokenize and remove stopwords

```{r}
stop_fr <- stopwords("fr", source = "snowball")

catalog_tokens <- catalog |>
  unnest_tokens(word, description) |>
  filter(
    !word %in% stop_fr,
    str_detect(word, "^\\p{L}+$")
  )

catalog_tokens |>
  count(word, sort = TRUE) |>
  slice_head(n = 12)
```

::: {.callout-tip collapse="true"}
Each row now represents a retained word. Frequencies offer a first view, but a description should be reread in context before interpretation.
:::

### Exercise 4 - Compare words by producer

```{r}
words_by_producer <- catalog_tokens |>
  count(producer, word, sort = TRUE) |>
  group_by(producer) |>
  slice_max(n, n = 5, with_ties = FALSE) |>
  ungroup()

words_by_producer
```

```{r}
ggplot(
  words_by_producer,
  aes(x = n, y = fct_reorder(word, n))
) +
  geom_col() +
  facet_wrap(vars(producer), scales = "free_y") +
  labs(
    x = "Occurrences",
    y = NULL,
    title = "Frequent words by producer"
  ) +
  theme_minimal()
```

::: {.callout-tip collapse="true"}
A difference may come from published topics, institutional vocabulary or description length. It does not directly measure data quality.
:::

## Block B - Build a simple lexical score

### Exercise 5 - Create a precision lexicon

Concrete terms receive `+1` and very general terms receive `-1`.

```{r}
precision_lexicon <- tibble(
  word = c(
    "localisation", "inventaire", "nombre", "mesure", "date",
    "horaire", "statistiques", "registre", "liste", "géographique",
    "quotidien", "annuel", "donnée", "données", "information",
    "informations", "ensemble", "divers", "autre", "différents"
  ),
  class = c(rep("concrete", 12), rep("general", 8)),
  score = if_else(class == "concrete", 1, -1)
)

precision_lexicon
```

::: {.callout-tip collapse="true"}
This score is not a validated measure of writing quality. It demonstrates lexicon scoring and must always be paired with human reading.
:::

### Exercise 6 - Calculate a score by record

```{r}
record_scores <- catalog_tokens |>
  inner_join(precision_lexicon, by = "word") |>
  group_by(dataset_id, producer, category) |>
  summarise(
    precision_score = sum(score),
    recognized_words = n(),
    .groups = "drop"
  ) |>
  right_join(
    catalog,
    by = c("dataset_id", "producer", "category")
  ) |>
  mutate(
    precision_score = replace_na(precision_score, 0),
    recognized_words = replace_na(recognized_words, 0),
    days_since_update = as.integer(snapshot_date - updated_at)
  )

record_scores |>
  select(
    dataset_id,
    producer,
    title,
    precision_score,
    recognized_words
  ) |>
  slice_head(n = 10)
```

::: {.callout-tip collapse="true"}
A zero may mean neutral vocabulary or no match in the small lexicon. `recognized_words` is therefore essential.
:::

### Exercise 7 - Summarize scores by producer

```{r}
precision_by_producer <- record_scores |>
  group_by(producer) |>
  summarise(
    mean_score = mean(precision_score),
    mean_recognized_words = mean(recognized_words),
    mean_resources = mean(resource_count),
    records = n(),
    .groups = "drop"
  )

precision_by_producer
```

```{r}
ggplot(
  precision_by_producer,
  aes(x = mean_score, y = reorder(producer, mean_score))
) +
  geom_col(fill = "steelblue") +
  labs(
    x = "Mean lexical score",
    y = NULL,
    title = "Exploratory lexical score by producer"
  ) +
  theme_minimal()
```

::: {.callout-tip collapse="true"}
The chart compares the sample, not overall producer quality. Topic and writing-style differences may explain part of the variation.
:::

## Block C - Identify distinctive words

### Exercise 8 - Calculate TF-IDF by producer

```{r}
producer_tfidf <- catalog_tokens |>
  count(producer, word) |>
  bind_tf_idf(word, producer, n) |>
  arrange(desc(tf_idf))

producer_tfidf |>
  group_by(producer) |>
  slice_max(tf_idf, n = 5, with_ties = FALSE) |>
  ungroup()
```

::: {.callout-tip collapse="true"}
A distinctive word is relatively more associated with one producer. It is not necessarily positive, negative or important to users.
:::

### Exercise 9 - Visualize distinctive words

```{r}
tfidf_top <- producer_tfidf |>
  group_by(producer) |>
  slice_max(tf_idf, n = 4, with_ties = FALSE) |>
  ungroup() |>
  mutate(word = fct_reorder(word, tf_idf))

ggplot(tfidf_top, aes(x = tf_idf, y = word)) +
  geom_col(fill = "darkorange") +
  facet_wrap(vars(producer), scales = "free_y") +
  labs(
    x = "TF-IDF",
    y = NULL,
    title = "Distinctive words in descriptions"
  ) +
  theme_minimal()
```

::: {.callout-tip collapse="true"}
Return to the original descriptions to check meaning before writing a conclusion.
:::

## Block D - Prepare dashboard output

### Exercise 10 - Build a summary table

```{r}
dashboard_summary <- record_scores |>
  group_by(producer) |>
  summarise(
    records = n(),
    mean_precision_score = mean(precision_score),
    mean_resources = mean(resource_count),
    median_days_since_update = median(days_since_update),
    categories = n_distinct(category),
    .groups = "drop"
  )

dashboard_summary
```

::: {.callout-tip collapse="true"}
This table separates indicator preparation from display. A dashboard could then present cards, charts and source descriptions.
:::

### Exercise 11 - Simulate a filter

```{r}
selected_producer <- "Ville de Québec"

dashboard_summary |>
  filter(producer == selected_producer)
```

::: {.callout-tip collapse="true"}
In a `shiny` app, this value could come from a menu. The analytical operation remains a filter on a prepared table.
:::

## Case Study 1 - Quality of dataset descriptions

A team wants to identify descriptions that deserve editorial review.

```{r}
records_to_review <- record_scores |>
  arrange(precision_score, recognized_words) |>
  select(
    producer,
    title,
    description,
    precision_score,
    recognized_words,
    days_since_update
  ) |>
  slice_head(n = 10)

records_to_review
```

::: {.callout-tip collapse="true"}
The score may prioritize reading, but should not automatically trigger correction. The lexicon is small, topics differ and a general description may still be accurate and useful.
:::

## Case Study 2 - Quebec tourism events

Import `quebec_tourism_events.csv`. `text` combines public SIT Québec fields to create a short reproducible corpus.

```{r}
events <- read_csv(
  "data/quebec_tourism_events.csv",
  show_col_types = FALSE
)

event_tokens <- events |>
  unnest_tokens(word, text) |>
  filter(
    !word %in% stop_fr,
    str_detect(word, "^\\p{L}+$")
  )

event_tfidf <- event_tokens |>
  count(event_type, word) |>
  bind_tf_idf(word, event_type, n) |>
  group_by(event_type) |>
  slice_max(tf_idf, n = 4, with_ties = FALSE) |>
  ungroup()

event_tfidf
```

```{r}
event_summary <- events |>
  group_by(tourism_region) |>
  summarise(
    events = n(),
    municipalities = n_distinct(municipality),
    median_duration_days = median(duration_days),
    proportion_with_website = mean(website_available),
    .groups = "drop"
  ) |>
  arrange(desc(events), tourism_region)

event_summary
```

::: {.callout-tip collapse="true"}
Useful visuals include a TF-IDF chart by event type and a regional coverage chart. The sample is balanced to at most eight events per tourism region, so it does not measure the true volume of the entire tourism offering.
:::
