---
title: "Adventure 8 - Open Data from Quebec"
subtitle: "STT-1100 Introduction to Data Science"
author:
  - institute: "Université Laval"
editor: visual
editor_options:
  chunk_output_type: console
format:
  html:
    toc: true
    toc-title: Plan
    css: [../css/base_css.css]
    embed-resources: true
---

```{r setup, include=FALSE}
library(tidyverse)
library(rvest)
library(purrr)

fallback_dataset_html <- '
<html>
  <body>
    <div class="dataset-content">
      <h3 class="dataset-heading"><a>Example public dataset</a></h3>
      <div class="dqc-org-cat">Organisation : Données Québec</div>
      <div class="dqc-org-cat">Catégories : Santé</div>
    </div>
  </body>
</html>'

safe_read_lines <- function(url, fallback = character()) {
  tryCatch(
    readLines(url, warn = FALSE),
    error = function(error) {
      message("Unable to download ", url, "; using a local fallback.")
      fallback
    }
  )
}

safe_read_html <- function(url) {
  tryCatch(
    read_html(url),
    error = function(error) {
      message("Unable to download ", url, "; using a local HTML fallback.")
      read_html(fallback_dataset_html)
    }
  )
}

extract_value <- function(text, pattern) {
  value <- text[grepl(pattern, text, ignore.case = TRUE)][1]
  if (length(value) == 0 || is.na(value)) {
    return(NA_character_)
  }
  trimws(sub(pattern, "", value, ignore.case = TRUE))
}
```

# Scenario

You work as an independent data science consultant. Unlike a salaried position, consulting work is organized around specific client mandates. You must be autonomous, rigorous and able to deliver effective, reusable and well-documented solutions.

Today, you are hired by Marie-Pier, research director at the Quebec Institute for Sustainable Data. She wants to explore the [Données Québec](https://www.donneesquebec.ca) portal to identify recent and relevant datasets. Your mission is to design a reliable scraping tool that extracts dataset titles, producers and categories, while respecting good technical and ethical practices.

::: callout-note
Your contact: Marie-Pier

She accompanies you throughout this adventure. She asks key questions, checks your results and helps you clarify your deliverables.
:::

<!-- adventure-brief:start -->
<div class="adventure-brief">
<p class="adventure-brief-title">Character card</p>
<div class="adventure-brief-grid">
<div class="adventure-brief-item">
<span class="adventure-brief-label">Your role</span>
<span class="adventure-brief-value">Independent data science consultant</span>
</div>
<div class="adventure-brief-item">
<span class="adventure-brief-label">Main contact</span>
<span class="adventure-brief-value">Marie-Pier</span>
</div>
<div class="adventure-brief-item">
<span class="adventure-brief-label">Organization and context</span>
<span class="adventure-brief-value">Quebec Institute for Sustainable Data</span>
</div>
<div class="adventure-brief-item">
<span class="adventure-brief-label">Mission</span>
<span class="adventure-brief-value">Create a robust, testable and respectful scraping tool</span>
</div>
<div class="adventure-brief-item">
<span class="adventure-brief-label">Data</span>
<span class="adventure-brief-value">Données Québec page or small local HTML test page</span>
</div>
<div class="adventure-brief-item">
<span class="adventure-brief-label">Deliverable</span>
<span class="adventure-brief-value"><code>IDUL.R</code> with <code>scrape_page()</code></span>
</div>
</div>
</div>
<!-- adventure-brief:end -->

## Adventure Objectives

- Understand the basics of web scraping with `rvest`.
- Create a function to extract metadata from a web page.
- Automate extraction over several pages.
- Explore trends in Quebec open data.
- Evaluate the ethical and technical limits of automated collection.

<!-- week-new:start -->
::: {.callout-note title="What is new this week"}
- You read the structure of a web page as a data source.
- You turn a manual extraction into a reusable function.
- You test your function on a local page to remain independent of portal changes.
- You connect code to ethical limits: `robots.txt`, terms of use and server load.
:::

<!-- week-new:end -->

# Before Scraping: Check the Context

::: {.callout-important title="Question from Marie-Pier"}
*What can we extract from this portal? Is scraping reasonable here?*
:::

A first reflex is to consult the site's `robots.txt` file. This file indicates the paths a site asks robots to avoid. It does not replace terms of use, ethical judgment or respect for server load, but it gives a useful signal before automating collection.

```{r exploration-ethics}
robots <- safe_read_lines(
  "https://www.donneesquebec.ca/robots.txt",
  fallback = c(
    "User-agent: *",
    "Disallow: /api/",
    "Disallow: /dataset/rate/"
  )
)

cat("Contents of robots.txt:\n")
writeLines(robots)

disinstructions <- robots[grepl("^Disallow", robots)]

cat("\nPaths prohibited for robots:\n")
writeLines(disinstructions)
```

::: callout-warning
A `robots.txt` file that does not explicitly forbid a page does not mean that everything is automatically acceptable. In this module, we limit collection to a few public result pages, add a pause between requests and do not bypass protection mechanisms.
:::

# Understanding Web Scraping with `rvest`

In this section, you discover the essential functions of the `rvest` package. The goal is to extract structured HTML content from a search page.

```{r reading-html}
url <- "https://www.donneesquebec.ca/recherche/?sort=metadata_modified+desc&page=1"
page <- safe_read_html(url)
page
```

The `read_html()` function downloads and converts the web page so it can be manipulated.

::: {.callout-important title="Question from Marie-Pier"}
*Can you show me what the structure of this page looks like? Can you spot an interesting element?*
:::

We can now target HTML elements with `html_elements()`:

```{r test-nodes}
blocks <- html_elements(page, ".dataset-content")
length(blocks)
blocks[[1]]
```

To extract text from an HTML element:

```{r test-text}
html_text2(blocks[[1]])
```

Now, let us test extraction of the title:

```{r test-title}
html_elements(blocks[[1]], ".dataset-heading a") |>
  html_text2()
```

And for the producer? We need to locate a substructure containing the information:

```{r test-producer}
info <- html_elements(blocks[[1]], ".dqc-org-cat") |>
  html_text2()

info
```

The Données Québec page used here returns French labels. A robust function should therefore handle labels such as `Organisation`, `Catégorie` and `Catégories`, even when you are reading the English version of the course.

```{r extract-producer}
producer <- extract_value(
  info,
  "^(Organisation|Organisme|Producteur|Producer|Organization)\\s*:?\\s*"
)

producer
```

::: callout-tip
Why use `map_chr()`?

`map_chr()` belongs to the `purrr` package, which is part of the `tidyverse`. This function applies a function to each element of a list and returns a character vector. This is exactly what we need when we want one text value per HTML block.

``` r
producteurs <- map_chr(blocks, function(block) {
  info <- html_elements(block, ".dqc-org-cat") |>
    html_text2()

  extract_value(
    info,
    "^(Organisation|Organisme|Producteur|Producer|Organization)\\s*:?\\s*"
  )
})
```
:::

::: callout-tip
Your turn: from this same HTML block, find how to extract the categories associated with each dataset. The observed label may be `Catégorie`, `Catégories`, `Category` or `Categories`.
:::

# Guided Creation of the `scrape_page()` Function

Marie-Pier wants you to create a reusable function named `scrape_page()`. It takes a URL as input and returns a `data.frame` with exactly the following columns:

- `titre`;
- `producteur`;
- `categorie`.

The column names remain in French because the automatic tests will check this exact contract.

::: callout-note
The template repository contains a small local HTML page so the function can be tested even if the portal changes or the network is unavailable. Your function must therefore work with both a public page and this local source, without changing the output contract.
:::

Here is a skeleton to complete:

```{r, eval=FALSE}
scrape_page <- function(url) {
  page <- rvest::read_html(url)

  blocks <- rvest::html_elements(page, ".dataset-content")

  titres <- rvest::html_elements(blocks, ".dataset-heading a") |>
    rvest::html_text2()

  producteurs <- purrr::map_chr(blocks, function(block) {
    info <- rvest::html_elements(block, ".dqc-org-cat") |>
      rvest::html_text2()

    # Extract the organization or producer.
  })

  categories <- purrr::map_chr(blocks, function(block) {
    info <- rvest::html_elements(block, ".dqc-org-cat") |>
      rvest::html_text2()

    # Extract the category or categories.
  })

  data.frame(
    titre = titres,
    producteur = producteurs,
    categorie = categories,
    stringsAsFactors = FALSE
  )
}
```

::: callout-note
Test your function with page 3. It should return 20 rows and exactly the columns `titre`, `producteur`, `categorie`.
:::

# Manual Repetition, Then a `for` Loop

Now test the extraction of the first five pages by calling your function several times:

```{r, eval=FALSE}
p1 <- scrape_page("https://www.donneesquebec.ca/recherche/?sort=metadata_modified+desc&page=1")
p2 <- scrape_page("https://www.donneesquebec.ca/recherche/?sort=metadata_modified+desc&page=2")
```

::: callout-tip
Marie-Pier: *Have you noticed what changes in the URL each time? Can you generalize this behavior?*
:::

Reminder from module 1: a loop automates repetitive behavior.

```{r, eval=FALSE}
results <- data.frame()

for (i in 1:5) {
  url <- paste0(
    "https://www.donneesquebec.ca/recherche/?sort=metadata_modified+desc&page=",
    i
  )

  cat("Page", i, "in progress...\n")
  page_data <- scrape_page(url)

  results <- bind_rows(results, page_data)
  Sys.sleep(1)
}
```

# Exploration Guided by Marie-Pier

Use the `results` table to answer your client's questions. She expects precise and careful results.

::: {.callout-important title="Questions from Marie-Pier"}
1. What are the most common dataset categories?
2. Which organizations publish the most?
3. Do we observe a diversity of fields or a concentration around a few themes?
:::

# Ethical Reflection

::: {.callout-note title="For your working notes"}
Briefly report:

- what the `robots.txt` file indicates;
- the limits of this check;
- the behaviors you adopted to remain respectful, such as limiting the number of pages, adding a pause and not bypassing protection;
- the situation for two other websites, for example `bonjourquebec.com` or `ulaval.ca`, using `robots.txt` or terms of use.
:::

# Challenge to Submit

The challenge associated with this adventure is described on the page [Challenge 8 - Scraping Function](defi.qmd).

You must submit an `IDUL.R` file containing your `scrape_page()` function. The starter repository is the GitHub template `STT-1100/aventure-8`.

<!-- portfolio-trace:start -->
<div class="portfolio-trace">
<p class="portfolio-trace-title">Portfolio trace</p>
<p>Keep a trace proving that your function is robust and responsible.</p>
<ul>
<li>the output contract for <code>scrape_page()</code>;</li>
<li>one successful test on a local HTML page;</li>
<li>an ethics note about <code>robots.txt</code>, server load and terms of use;</li>
<li>a short exploration of the results, if a public page is available.</li>
</ul>
</div>
<!-- portfolio-trace:end -->

# Conclusion of the Adventure

You have designed a functional and automatable scraping tool, used a `for` loop, extracted structured metadata and deepened your understanding of scraping ethics.
