---
title: "Module 8 - Exercises"
subtitle: "STT-1100 Introduction to Data Science"
format:
  html:
    toc: true
    toc-title: Plan
    embed-resources: true
    link-external-newwindow: true
    css: [../css/base_css.css]
editor: visual
---

::: {.callout-important}
## Module workspace

<a class="btn btn-primary" href="../../downloads/donnees/stt1100-module-08-en.zip" download>Download the ready-to-open workspace (.zip)</a>

Unpack the file, then open the `.Rproj` project. The paths used in the exercises will work without moving the data.
:::


## Consolidation Exercises

These exercises are independent from the adventure and the challenge. They consolidate the technical moves of module 8: reading an HTML page, targeting elements with CSS selectors, turning an extraction into a function, automating repeated extraction and documenting collection limits.

The HTML pages used here are local snapshots of real Quebec data. Both catalogs come from the [official Données Québec API](https://www.donneesquebec.ca/page-api/){target="_blank"}; the events page comes from the Ministry of Tourism dataset [Événements - Système d'information touristique Québec](https://www.donneesquebec.ca/recherche/dataset/sit-quebec-evenements){target="_blank"}. The snapshots preserve the published values, while simplifying the HTML structure to keep the exercise stable and reproducible. The metadata and events are distributed under the CC BY 4.0 licence.

```{r setup, warning=FALSE, message=FALSE}
library(tidyverse)
library(rvest)
library(purrr)
```

## Block A - Read a Local HTML Page

### Exercise 1 - Import an HTML Page

Read `catalogue_donnees_quebec.html` with `read_html()`. Locate the six dataset cards with the `.dataset-card` selector. They cover Quebec City, Gatineau, Sherbrooke, Trois-Rivières, Saguenay and Montreal.

::: {.callout-note collapse="true"}
### Solution

```{r}
catalog_page <- read_html("data/catalogue_donnees_quebec.html")

cards <- catalog_page |>
  html_elements(".dataset-card")

length(cards)
cards[[1]]
```
:::

### Exercise 2 - Extract a First Vector

Extract dataset titles with the `.dataset-title` selector.

::: {.callout-note collapse="true"}
### Solution

```{r}
titles <- cards |>
  html_element(".dataset-title") |>
  html_text2()

titles
```
:::

### Exercise 3 - Build a Table

Extract producers, categories and update dates. Combine them into a tibble.

::: {.callout-note collapse="true"}
### Solution

```{r}
catalog <- tibble(
  title = html_element(cards, ".dataset-title") |> html_text2(),
  producer = html_element(cards, ".dataset-producer") |> html_text2(),
  category = html_element(cards, ".dataset-category") |> html_text2(),
  updated_at = html_element(cards, ".dataset-updated") |>
    html_text2() |>
    as.Date()
)

catalog
```
:::

## Block B - Turn Extraction into a Function

### Exercise 4 - Create a Helper Function

Create an `extract_text()` function that receives an HTML node and a CSS selector, then returns the corresponding text. If the element is missing, the function must return `NA_character_`.

::: {.callout-note collapse="true"}
### Solution

```{r}
extract_text <- function(node, selector) {
  element <- html_element(node, selector)
  text <- html_text2(element)

  if (length(text) == 0 || is.na(text) || !nzchar(text)) {
    return(NA_character_)
  }

  text
}

extract_text(cards[[1]], ".dataset-title")
extract_text(cards[[1]], ".missing-field")
```
:::

### Exercise 5 - Create an Extraction Function

Write an `extract_catalog(file)` function that reads a local HTML page and returns a tibble with the columns `title`, `producer`, `category`, `updated_at`.

::: {.callout-note collapse="true"}
### Solution

```{r}
extract_catalog <- function(file) {
  page <- read_html(file)
  cards <- html_elements(page, ".dataset-card")

  tibble(
    title = map_chr(cards, extract_text, ".dataset-title"),
    producer = map_chr(cards, extract_text, ".dataset-producer"),
    category = map_chr(cards, extract_text, ".dataset-category"),
    updated_at = map_chr(cards, extract_text, ".dataset-updated") |>
      as.Date()
  )
}

extract_catalog("data/catalogue_donnees_quebec.html")
```
:::

### Exercise 6 - Test an Irregular Page

Use the same function on `catalogue_donnees_quebec_irregulier.html`. Some official records have no category assigned in the portal. Which values are missing?

::: {.callout-note collapse="true"}
### Solution

```{r}
irregular_catalog <- extract_catalog("data/catalogue_donnees_quebec_irregulier.html")

irregular_catalog

irregular_catalog |>
  summarise(
    missing_producers = sum(is.na(producer)),
    missing_categories = sum(is.na(category))
  )
```

The function does not stop when a field is missing. It returns `NA`, which makes the problem diagnosable in a table.
:::

### Exercise 7 - Automate Across Several Pages

Use `purrr::imap_dfr()` to apply `extract_catalog()` to both catalog pages and add a `source` column.

::: {.callout-note collapse="true"}
### Solution

```{r}
catalog_files <- c(
  regular = "data/catalogue_donnees_quebec.html",
  irregular = "data/catalogue_donnees_quebec_irregulier.html"
)

combined_catalogs <- imap_dfr(
  catalog_files,
  function(file, source) {
    extract_catalog(file) |>
      mutate(source = source, .before = 1)
  }
)

combined_catalogs
```
:::

## Block C - Collection Sobriety and Ethics

### Exercise 8 - Interpret a `robots.txt`

Here is a fictitious `robots.txt` example.

```{r}
robots_example <- c(
  "User-agent: *",
  "Disallow: /admin/",
  "Disallow: /api/private/",
  "Crawl-delay: 1"
)

writeLines(robots_example)
```

Identify paths to avoid and explain why this file is not full permission to collect data.

::: {.callout-note collapse="true"}
### Solution

```{r}
disallowed_paths <- robots_example |>
  str_subset("^Disallow:") |>
  str_remove("^Disallow:\\s*")

disallowed_paths
```

`robots.txt` provides technical instructions to crawlers. It does not replace terms of use, ethical judgment, caution about server load or written permission when collection and redistribution are sensitive.
:::

## Case Study 1 - Quebec Municipal Catalog

You must produce a short summary of metadata published by several Quebec municipalities in Données Québec.

### Exercise 9 - Summarize Categories

Starting from the two combined catalog pages, compute the number of datasets by category.

::: {.callout-note collapse="true"}
### Solution

```{r}
category_summary <- combined_catalogs |>
  mutate(category = replace_na(category, "Not indicated")) |>
  count(category, sort = TRUE)

category_summary
```
:::

### Exercise 10 - Write a Collection Note

Write three sentences explaining what your code collects, how it limits risks and what should be checked before applying it to a real website.

::: {.callout-note collapse="true"}
### Sample Answer

The code reads only local HTML pages prepared for the exercise and extracts aggregated public fields. In a real context, it would be necessary to limit the number of pages, add pauses between requests and avoid mass collection or bypassing protections. It would also be necessary to check `robots.txt`, terms of use, data licenses and redistribution rules before publishing a derived table.
:::

## Case Study 2 - Public Events Page

The Ministry of Tourism publishes SIT Québec events. The snapshot retains six announced events in Gatineau, Montreal, Quebec City, Saguenay, Sherbrooke and Trois-Rivières. You want to extract titles, dates, locations and types to produce a calendar.

### Exercise 11 - Extract Events

Create an `extract_events(file)` function that returns a tibble with the columns `title`, `date`, `location`, `theme`.

::: {.callout-note collapse="true"}
### Solution

```{r}
extract_events <- function(file) {
  page <- read_html(file)
  events <- html_elements(page, ".event-card")

  tibble(
    title = map_chr(events, extract_text, ".event-title"),
    date = html_element(events, ".event-date") |>
      html_attr("datetime") |>
      as.Date(),
    location = map_chr(events, extract_text, ".event-location"),
    theme = map_chr(events, extract_text, ".event-theme")
  )
}

events <- extract_events("data/evenements_sit_quebec.html")

events
```
:::

### Exercise 12 - Check the Result

Produce a summary of the number of events by theme and write one simple check that could be used as a test.

::: {.callout-note collapse="true"}
### Solution

```{r}
events |>
  count(theme, sort = TRUE)

stopifnot(
  nrow(events) == 6,
  identical(names(events), c("title", "date", "location", "theme")),
  all(!is.na(events$title))
)
```

These tests are simple, but they already detect common errors: wrong row count, incorrectly named columns or titles that were not extracted.
:::
