---
title: "Module 7 - Exercises"
subtitle: "STT-1100 Introduction to Data Science"
format:
  html:
    toc: true
    toc-title: Plan
    embed-resources: true
    link-external-newwindow: true
    css: [../css/base_css.css]
editor: visual
---

::: {.callout-important}
## Module workspace

<a class="btn btn-primary" href="../../downloads/donnees/stt1100-module-07-en.zip" download>Download the ready-to-open workspace (.zip)</a>

Unpack the file, then open the `.Rproj` project. The paths used in the exercises will work without moving the data.
:::


## Consolidation Exercises

These exercises are independent from the adventure and challenge. They consolidate the Module 7 skills: recognizing misleading visualizations, rebuilding defensible charts, reducing re-identification risks and writing short ethical notes.

::: {.callout-note}
## Real data that are already aggregated

The exercises use [Sherbrooke public-safety incidents](https://www.donneesquebec.ca/recherche/dataset/64d19d62f0804f5896e4b24c32aea49d_0), [Statistics Canada annual population estimates](https://www150.statcan.gc.ca/t1/tbl1/en/tv.action?pid=1710015501) and the [Données Québec user survey](https://www.donneesquebec.ca/recherche/dataset/sondage). The script `scripts/build_module07_real_data.R` removes coordinates and exact incident dates, then aggregates observations by month and type. The survey is originally published as aggregate counts and percentages, without individual responses.
:::

```{r setup, warning=FALSE, message=FALSE}
library(tidyverse)
library(scales)
```

## Block A - Diagnose a Visualization

### Exercise 1 - Compare Two Axes

Import the aggregated Sherbrooke incidents, then produce two charts of the same totals by type:

- one chart with a truncated vertical axis;
- one chart whose axis starts at zero.

How does visual perception change?

::: {.callout-note collapse="true"}
### Solution

```{r}
incidents <- read_csv(
  "data/incidents_securite_sherbrooke_agreges.csv",
  show_col_types = FALSE
)

incidents_type <- incidents |>
  group_by(type_incident) |>
  summarise(
    total_incidents = sum(nb_incidents),
    .groups = "drop"
  ) |>
  arrange(desc(total_incidents))

incidents_type
```

```{r}
#| fig-alt: "Misleading horizontal bar chart of incidents by type, with a truncated axis that exaggerates differences."
ggplot(
  incidents_type,
  aes(x = reorder(type_incident, total_incidents), y = total_incidents)
) +
  geom_col(fill = "#B33A3A") +
  coord_flip(ylim = c(1000, max(incidents_type$total_incidents) * 1.05)) +
  labs(
    title = "Misleading version: truncated axis",
    x = "Incident type",
    y = "Number of incidents"
  )
```

```{r}
#| fig-alt: "Defensible horizontal bar chart of incidents by type, with an axis starting at zero."
ggplot(
  incidents_type,
  aes(x = reorder(type_incident, total_incidents), y = total_incidents)
) +
  geom_col(fill = "#3C6E71") +
  coord_flip() +
  scale_y_continuous(
    limits = c(0, NA),
    expand = expansion(mult = c(0, 0.05))
  ) +
  labs(
    title = "Defensible version: axis starts at zero",
    x = "Incident type",
    y = "Number of incidents"
  )
```

The truncated axis magnifies differences and even makes some small categories invisible. Bar length encodes quantity, so the axis should normally start at zero.
:::

### Exercise 2 - Publish Aggregate Counts

Produce a chart showing totals by type, counts and a note defining the scope of the data.

::: {.callout-note collapse="true"}
### Solution

```{r}
#| fig-alt: "Horizontal bar chart of public-safety incidents by type, with counts printed on the bars."
ggplot(
  incidents_type,
  aes(x = reorder(type_incident, total_incidents), y = total_incidents)
) +
  geom_col(fill = "#4C78A8") +
  geom_text(aes(label = comma(total_incidents)), hjust = -0.1, size = 3.4) +
  coord_flip() +
  scale_y_continuous(
    limits = c(0, max(incidents_type$total_incidents) * 1.15)
  ) +
  labs(
    title = "Public-safety incidents by type",
    subtitle = "Sherbrooke, 2022 to 2024; aggregated monthly data",
    x = NULL,
    y = "Number of incidents"
  )
```

This chart publishes no address, coordinate or exact date. It describes incidents recorded in the source, not every situation that may have occurred.
:::

### Exercise 3 - Avoid an Unfair Comparison Between Years

Join annual totals to population estimates and calculate a rate per 10,000 people. Compare this rate with raw counts.

::: {.callout-note collapse="true"}
### Solution

```{r}
population <- read_csv(
  "data/population_sherbrooke_2022_2024.csv",
  show_col_types = FALSE
)

incidents_year <- incidents |>
  group_by(annee) |>
  summarise(
    total_incidents = sum(nb_incidents),
    .groups = "drop"
  ) |>
  left_join(population, by = "annee") |>
  mutate(
    rate_per_10000 = total_incidents / population_estimee * 10000
  )

incidents_year
```

The rate accounts for population change. Years are still not perfectly comparable if recording practices or context changed.
:::

## Block B - Reduce Re-identification Risks

### Exercise 4 - Identify Risky Variables

The original source includes `OBJECTID`, `DATEINCIDENT`, `DATETEXTE`, `x`, `y`, `DESCRIPTION` and `ANNEE`. Classify these fields by disclosure risk and compare them with the course extract.

::: {.callout-note collapse="true"}
### Solution

```{r}
tibble(
  source_variable = c(
    "OBJECTID", "DATEINCIDENT", "DATETEXTE", "x", "y",
    "DESCRIPTION", "ANNEE"
  ),
  assessment = c(
    "technical identifier removed",
    "exact date and time removed",
    "exact date removed",
    "coordinate removed",
    "coordinate removed",
    "category retained",
    "broad period retained"
  )
)

names(incidents)
```

Monthly aggregation reduces temporal precision and removing coordinates eliminates fine location. This reduces risk without guaranteeing absolute anonymization.
:::

### Exercise 5 - Apply a Publication Threshold

Create `public_count`, masking monthly cells containing fewer than five incidents.

::: {.callout-note collapse="true"}
### Solution

```{r}
public_incidents <- incidents |>
  mutate(
    public_count = if_else(nb_incidents < 5, NA_integer_, nb_incidents),
    publication_status = if_else(
      is.na(public_count),
      "suppressed: small count",
      "published"
    )
  )

public_incidents |>
  filter(is.na(public_count))
```

Five is a cautious teaching rule, not a universal anonymity guarantee. A real threshold depends on context, other variables and organizational rules.
:::

### Exercise 6 - Measure the Effect of Suppression

Count published and suppressed cells by incident type.

::: {.callout-note collapse="true"}
### Solution

```{r}
public_incidents |>
  count(type_incident, publication_status) |>
  pivot_wider(
    names_from = publication_status,
    values_from = n,
    values_fill = 0
  )
```
:::

### Exercise 7 - Visualize Without Overexposing Small Categories

Calculate annual totals by type, retain only cells with at least 20 incidents and graph rates per 10,000 people.

::: {.callout-note collapse="true"}
### Solution

```{r}
#| fig-alt: "Grouped bar chart of recorded incident rates by year and type, for categories containing at least twenty incidents."
rates_type_year <- incidents |>
  group_by(annee, type_incident) |>
  summarise(
    nb_incidents = sum(nb_incidents),
    .groups = "drop"
  ) |>
  left_join(population, by = "annee") |>
  filter(nb_incidents >= 20) |>
  mutate(
    rate_per_10000 = nb_incidents / population_estimee * 10000
  )

ggplot(
  rates_type_year,
  aes(x = factor(annee), y = rate_per_10000, fill = type_incident)
) +
  geom_col(position = "dodge") +
  labs(
    title = "Recorded incident rates by type",
    subtitle = "Annual categories with 20 incidents or more",
    x = "Year",
    y = "Incidents per 10,000 people",
    fill = "Type"
  )
```

The graph should not be used to label a neighbourhood or population. It presents city-wide recorded categories.
:::

## Case Study 1 - Responsible Municipal Publication

### Exercise 8 - Build a Publishable Table

Using `public_incidents`, create a monthly table retaining type, year and month but not publishing small numbers.

::: {.callout-note collapse="true"}
### Solution

```{r}
publication_table <- public_incidents |>
  select(
    annee,
    mois,
    type_incident,
    public_count,
    publication_status
  ) |>
  arrange(annee, mois, type_incident)

publication_table
```
:::

### Exercise 9 - Write a Short Ethical Note

Write three sentences explaining what is published, what is suppressed and the main interpretation limitation.

::: {.callout-note collapse="true"}
### Example Answer

The publication presents monthly aggregate counts by type, without exact dates or coordinates. Cells with fewer than five incidents are suppressed to reduce disclosure risk and overinterpretation of very small numbers. These data describe recorded incidents and measure neither every situation that occurred nor the causes of observed variation.
:::

## Case Study 2 - Public Survey and Cautious Communication

The file `data/sondage_utilisateurs_donnees_quebec_2020_2025.csv` contains aggregate responses to three questions from the Données Québec public consultation. It contains neither respondent identifiers nor free-text comments.

### Exercise 10 - Inspect Counts and Missing Values

Import the file, count missing values and summarize recorded selections by year and question.

::: {.callout-note collapse="true"}
### Solution

```{r}
survey <- read_csv(
  "data/sondage_utilisateurs_donnees_quebec_2020_2025.csv",
  show_col_types = FALSE
)

survey |>
  summarise(
    missing_counts = sum(is.na(nb_reponses)),
    missing_percentages = sum(is.na(pourcentage))
  )

survey_summary <- survey |>
  filter(!is.na(nb_reponses)) |>
  group_by(annee, question) |>
  summarise(
    total_selections = sum(nb_reponses),
    response_categories = n(),
    .groups = "drop"
  )

survey_summary
```

The total number of selections is not necessarily the number of people when a question permits multiple responses.
:::

### Exercise 11 - Produce a Defensible Chart

For question 6 in 2025, suppress response categories with fewer than five selections and produce a chart with counts.

::: {.callout-note collapse="true"}
### Solution

```{r}
#| fig-alt: "Horizontal bar chart of responses to question 6 in 2025 after hiding categories with fewer than five selections."
survey_2025_q6 <- survey |>
  filter(
    annee == 2025,
    question == 6,
    !is.na(nb_reponses),
    nb_reponses >= 5
  )

ggplot(
  survey_2025_q6,
  aes(x = fct_reorder(reponse, nb_reponses), y = nb_reponses)
) +
  geom_col(fill = "#2F4B7C") +
  geom_text(aes(label = nb_reponses), hjust = -0.1, size = 3.4) +
  coord_flip() +
  scale_y_continuous(
    limits = c(0, max(survey_2025_q6$nb_reponses) * 1.2)
  ) +
  labs(
    title = "Reported uses of open data",
    subtitle = "Question 6, 2025; categories below five selections omitted",
    x = NULL,
    y = "Number of selections"
  )
```

Participants chose to take part in an online consultation. Results should not be presented as representative of the entire Québec population.
:::

### Exercise 12 - Decide What Should Not Be Published

Name two types of information that should not be added to this open publication and explain why.

::: {.callout-note collapse="true"}
### Example Answer

An identifier linking selections from the same person should not be added because it would turn an aggregate table into an individual history. Free-text comments should not be published without a rigorous process because they may contain names, organizations, contact details or recognizable situations. Even in an aggregate table, very small counts require review before release.
:::
